---
title: "bulk-rnaseq_setbp1"
author: "Gokberk Alagoz, Maggie Wong, L&G Dept., MPI for Psycholinguistics"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rsubread)
library(biomaRt)
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DESeq2)
library(vsn)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(topGO)
library(VennDiagram)
library("ggbeeswarm")
library(EnsDb.Hsapiens.v79)
library(RRHO)
library(RRHO2)
library(umap)
library(rrvgo) #reduce redundant GO term
library(gridExtra)
library(stats)  # for Fisher's exact test
```

## Rsubread and data preparation
this part is for only plotting PCA including d0 samples
```{r data_prep}
#deseq_outDir = "/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/"

if (Sys.info()['sysname']=='Windows') {dir="P://workspaces/"} else {dir="/data/workspaces/lag/workspaces/"}
#deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_allsamples_MW20250106/",sep="")
#deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_noOutlier_MW20250106/",sep="")
deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_noOutlier_noD0_MW20250107/1_condition/",sep="")

#always save RData
#save.image(file = paste0(dir, "lg-scrna/working_data/Maggie/working/script/bulk_rna_SETBP1_iNeuron/3_featureCounts_deseq2/setbp1_iN_rna_d10-12condition.RData"))

# load the image
load(file = paste0(dir, "lg-scrna/working_data/Maggie/working/script/bulk_rna_SETBP1_iNeuron/3_featureCounts_deseq2/setbp1_iN_rna_d10-12condition.RData"))

# Read bam files
list_of_bam1 = list.files(path = "/data/workspaces/lag/workspaces/lg-scrna/primary_data/maggie_data_release/bulk_rna_SETBP1_iNeuron/X204SC24102861-Z01-F002_01/02.Bam", pattern = ".bam$", full.names = T)
list_of_bam2 = list.files(path = "/data/workspaces/lag/workspaces/lg-scrna/primary_data/maggie_data_release/bulk_rna_SETBP1_iNeuron/X204SC24102861-Z01-F002_04/02.Bam", pattern = ".bam$", full.names = T)
list_of_bam3 = list.files(path = "/data/workspaces/lag/workspaces/lg-scrna/primary_data/maggie_data_release/bulk_rna_SETBP1_iNeuron/X204SC24102861-Z01-F002_05/02.Bam", pattern = ".bam$", full.names = T)

list_of_bam = c(list_of_bam1,list_of_bam2,list_of_bam3)

# Create a count table
rsubread.counts = featureCounts(list_of_bam,
                               #annot.ext="/data/workspaces/lag/workspaces/lg-scrna/primary_data/maggie_data_release/20211123_setbp1_hdf_b2_rna/X201SC21072043-Z01-F002/ref/genome.gtf/genome.gtf",
                               annot.ext="/data/workspaces/lag/workspaces/lg-scrna/primary_data/maggie_data_release/bulk_rna_SETBP1_iNeuron/X204SC24102861-Z01-F002_01/04.Ref/genome.gtf",
                               isGTFAnnotationFile = TRUE,
                               GTF.featureType = "exon",
                               GTF.attrType = "gene_id",
                               useMetaFeatures = TRUE,
                               allowMultiOverlap = FALSE,
                               isPairedEnd = TRUE,
                               requireBothEndsMapped = TRUE)

# Save the count table as an R object
save(rsubread.counts, file = "/data/workspaces/lag/workspaces/lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/3_featureCounts/rsubread.counts.RData")


load(file = paste0(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/3_featureCounts/rsubread.counts.RData"))
################################################

# Get read counts per gene
rcounts = rsubread.counts$counts
# Get gene annotations
annot = rsubread.counts$annotation

head(rcounts)
dim(rcounts)
# there are 58,735 genes and 54 samples

# Load the metadata of your samples
inf = read.table(paste0(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/sample_name_condition_v1.csv"), skipNul = T, header = T, sep=",")


```

## prep for deseq2

```{r prep4deseq}

# Check the mean gene length
summary(annot$Length)

# Take out all transcripts that have no reads in any of the experiments
rcounts = rcounts[(rowSums(rcounts)>0),]

# Make countdata a dataframe
countdata = data.frame(GeneID = rownames(rcounts), rcounts)

# Add geneLength to annot
# Here, 150 in mutate() function is the mean read length. I pulled out this number from multiqc results:
# /data/workspaces/lag/workspaces/lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_HDF/multiqc/multiqc/Alignment_QC_Report_data.pdf
countdata = annot %>%
            dplyr::select(GeneID, Length) %>%
            inner_join(countdata) %>%
            mutate(Length = Length-150+1) %>%
            dplyr::filter(Length>1) # filter short genes

# Now we divide first by the gene Length and then the colsums
# This changes all numeric colums
# The dot refers to the current value
tpm = countdata %>%
      mutate_if(is.numeric,funs(rpk=./Length*1e3)) %>%
      mutate_if(is.numeric,funs(tpm=./sum(.)*1e6)) %>%
      # only want to keep the new columns
      dplyr::select(GeneID, contains("rpk_tpm"), -Length_rpk_tpm)

#change column names back
names(tpm) = colnames(countdata)[-2]

```

# Check how comparable the TPMs are across samples

```{r checkpoint_plot1}

# Avoid having log(0) by simply adding the minimum TPM > 0 to each sample

expMin = apply(tpm[,-1],2,function(x){min(x[x>0])})

log_tpm = log10(t(tpm[,-1]) + expMin)

# Transposing with t() made TPM a matrix,
# Transpose back and revert to data.frame
# Make long format

tpm_long = data.frame(t(log_tpm)) %>%
           gather(key= "SRR", value = "log10_TPM" )

head(tpm_long)

ggplot(tpm_long,aes(x=SRR,y=log10_TPM)) +
       geom_boxplot() +
       coord_flip() +
       ylab(expression(log[10](TPM)))

ggsave(paste0(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_allsamples_MW20250106/tpm_boxplots/TPM_comparison_boxplots_v1.pdf"), width = 12, height = 20, units = "cm")

```
B7D10_2089 has very low tpm, remove?

# Continue filtering before running DE analysis.

```{r prep4deseq_part2}

nrow(countdata)
# There are still 42861 genes in our count table
# Filter further to 1) get rid off zeros in the matrix and
# improve normalization and 2) to filter out very lowly
# expressed genes, cuz you can't detect differential
# expression for those genes anyway.

# Calculate how many genes are expressed in each sample
tpm %>% summarise_if(is.numeric, funs(sum(.>=1)))
number_genes_expressed = tpm %>% summarise_if(is.numeric, funs(sum(.>=1)))

write.csv(number_genes_expressed, file =paste0(deseq_outDir, "number_genes_expressed.csv"))

# Move geneIDs to rownames, remove that column
rownames(tpm) = tpm[,1]
tpm = tpm[,-1]

# Condition only 10 samples are allowed to have a TPM < 1
# Note that this 10 samples threshold is arbitrary.
cut = dim(tpm)[2] - 10
sum(apply(tpm, 1, function(x){sum(x>=1)>=cut}))
# 12563 genes pass this threshold
# So we reduced gene number from 42k to ~12k, success!
# is this too harsh?

# try another filter threshold
# Filter genes with TPM ≥ 1 in at least 10% of samples, less stringent filtering
cut2 <- 0.1 * ncol(tpm)  # 10% of samples
sum(apply(tpm, 1, function(x){sum(x>=1)>=cut2}))
# 18639 genes pass this threshold
# will too many zeros affect the deseq2 results? there are some very lowly expressed genes

# Condition only 20 samples are allowed to have a TPM < 1 so around half of the samples
cut3 = dim(tpm)[2] - 20
sum(apply(tpm, 1, function(x){sum(x>=1)>=cut3}))
# 13620 genes pass this threshold
# some less expressed genes are included but not too crazy

# first try to go with cut2
# do the actual filtering
filt = tpm[apply(tpm, 1, function(x){sum(x>=1)>=cut2}),]
tpm_filt_counts = rcounts[match(rownames(filt),rownames(rcounts)),]
write.csv(tpm_filt_counts, file =paste0(deseq_outDir, "tpm_filt_counts.csv"))

```
B7D10_2089 has 12700 genes expressed, similar to others
first try to go with cut2, 18639 genes pass this threshold

# Run differential expression analysis with "DESeq2".
## for all samples inclduing d0 for PCA
https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#introduction
```{r DEanalysis-part1-1}

head(inf)

# Make sure that the IDs in the info table have
# the same order as the columns in the count table
rownames(inf) = inf$ID # make ID rownames
colnames(rcounts) = str_replace(colnames(rcounts), ".bam", "") # remove ".bam" from colnames
colnames(tpm_filt_counts) = str_replace(colnames(tpm_filt_counts), ".bam", "") # remove ".bam" from colnames
tpm_filt_counts = tpm_filt_counts[, rownames(inf)] # order columns based on sample list in inf

colnames(tpm_filt_counts) == rownames(inf) # check if it worked (all should be TRUE, yes)
colnames(tpm_filt_counts) = inf$ID # make IDs in info as colnames of filtered counts, data still the same
rownames(inf) = inf$ID # make IDs back to rowname
colnames(tpm_filt_counts) == rownames(inf) # check if it worked (all should be TRUE, yes, hope everything is fine)

nrow(tpm_filt_counts)
#[1] 18639

inf$sex <- factor(inf$sex)
inf$batch <- factor(inf$batch)
inf$day <- factor(inf$day)
inf$condition <- factor(inf$condition)
inf$genotype <- factor(inf$genotype)

# designs

design1 = ~ sex + batch + day
design2= ~ sex + batch + day + condition
# d0 creates problem for day:condition so will first do design2 for the pca
inf_subset <- inf[inf$day != "d0", ]
design3 = ~ sex + batch + day + condition + condition:day

# DESeq design without time-course for plotting PCA only
dds = DESeqDataSetFromMatrix(countData = tpm_filt_counts,
                             colData = inf,
                             design = design2)

design(dds)
#~sex + batch + day + condition

dds = estimateSizeFactors(dds)
summary(sizeFactors(dds))
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.7025  0.9359  1.0069  1.0178  1.1045  1.4086 

#saveRDS(dds, file="/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/deseq_results/dds_v2.rds")
#readRDS(dds, file="/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/deseq_results/dds_v2.rds")
saveRDS(dds, file =paste0(deseq_outDir,"/deseq_results/dds_allsamples_v1.rds"))
#readRDS(dds, file =paste0(deseq_outDir,"/deseq_results/dds_allsamples_v1.rds"))

vsd = varianceStabilizingTransformation(dds)

# Remove batch effect only for visualization purpose.
# Important Note: DESeq2 takes raw count numbers as input, and
# models batch effect. Thus, DE gene lists are not affected
# by this removeBatchEffect step.
# See what Mike Love says here: https://support.bioconductor.org/p/104290/
# assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$batch)
# plotPCA(vsd, "batch")

# this is for removng batch effects and covariate effect
design <- model.matrix(~ sex + batch + age_at_sampling + day + condition, data = as.data.frame(colData(vsd)))
assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$batch, design = design)

# Extract a matrix of normalized counts from the S4 object vsd
vsdMat = assay(vsd)

pdf(paste0(deseq_outDir, "/SDplots/tpm.pdf"), width = 7, height = 8)
meanSdPlot(as.matrix(tpm))
dev.off()

pdf(paste0(deseq_outDir, "/SDplots/log_tpm.pdf"), width = 7, height = 8)
meanSdPlot(t(log_tpm))
dev.off()

pdf(paste0(deseq_outDir, "/SDplots/vsdMat.pdf"), width = 7, height = 8)
meanSdPlot(vsdMat)
dev.off()

# PCA using the 500 most variable genes from the variance-stabilized data
#nudge <- position_nudge(x = 2, y = 2)
desired_order_genotype <- c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13")
desired_order_condition <- c("Control", "Within_degron", "Outside_degron", "Truncating")
desired_order_ageAtSampling <- c("kid", "teenager", "adult")

vsd$genotype <- factor(vsd$genotype, levels = desired_order_genotype)
vsd$condition <- factor(vsd$condition, levels = desired_order_condition)
vsd$ageAtSampleing <- factor(vsd$condition, levels = desired_order_ageAtSampling )

#colour_genotype <- c("purple", "green", "blue", "orange")
#names(colour_genotype) <- levels(vsd$genotype)

p1 = plotPCA(vsd, intgroup = "condition", ntop = 500) +
  ggplot2::theme(legend.position = "top") + 
  geom_text_repel(aes(label = name), size = 2, max.overlaps = 50) +
  geom_point(size=0.1)


p2 = plotPCA(vsd, intgroup = "batch", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p3 = plotPCA(vsd, intgroup = "sex", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
    #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p4 = plotPCA(vsd, intgroup = "age_at_sampling", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p5 = plotPCA(vsd, intgroup = "day", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p6 = plotPCA(vsd, intgroup = "genotype", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1) #+ 
  #scale_colour_manual(values = colour_genotype)

pdf(paste0(deseq_outDir, "PCA/PCA_4var_allsamples.pdf"), width = 20, height = 30)
plot_grid(p1, p4, p3, p2, p5, p6, nrow = 3, ncol = 2)
dev.off()

```
3 samples were not clustering with the rest so going to remove them
B8D10_1223
B7D12_1225 
B7D12_1223
the last two seems to be swapped but to be safe, will remove both

```{r DEanalysis-part1-2}

head(inf)

# # Make sure that the IDs in the info table have
# # the same order as the columns in the count table
# rownames(inf) = inf$ID # make ID rownames
# colnames(rcounts) = str_replace(colnames(rcounts), ".bam", "") # remove ".bam" from colnames
# colnames(tpm_filt_counts) = str_replace(colnames(tpm_filt_counts), ".bam", "") # remove ".bam" from colnames
# tpm_filt_counts = tpm_filt_counts[, rownames(inf)] # order columns based on sample list in inf
# 
# colnames(tpm_filt_counts) == rownames(inf) # check if it worked (all should be TRUE, yes)
# colnames(tpm_filt_counts) = inf$ID # make IDs in info as colnames of filtered counts, data still the same
# rownames(inf) = inf$ID # make IDs back to rowname
# colnames(tpm_filt_counts) == rownames(inf) # check if it worked (all should be TRUE, yes, hope everything is fine)
# 
# nrow(tpm_filt_counts)
# #[1] 18639
# 
# inf$sex <- factor(inf$sex)
# inf$batch <- factor(inf$batch)
# inf$day <- factor(inf$day)
# inf$condition <- factor(inf$condition)
# inf$genotype <- factor(inf$genotype)

# designs
design1 = ~ sex + batch + day
design2= ~ sex + batch + day + condition
design3 = ~ sex + batch + day + condition + condition:day

# remove samples that are outlier
exclude_ids <- c("B8D10_1223", "B7D12_1225", "B7D12_1223")
inf_noOutlier <- inf[!inf$ID %in% exclude_ids, ]
#51 samples left

# # remove d0 for the rest of deseq2
# inf_noOutlier_noD0 <- inf_noOutlier[inf_noOutlier$day != "d0", ]

# remove outlier from the tpm_filt_counts
# Subset tpm_filt_counts to match the IDs in inf_subset
tpm_filt_counts_noOutlier <- tpm_filt_counts[, colnames(tpm_filt_counts) %in% inf_noOutlier$ID]

# Ensure the column order of tpm_filt_counts matches inf_subset
tpm_filt_counts_noOutlier <- tpm_filt_counts_noOutlier[, match(inf_noOutlier$ID, colnames(tpm_filt_counts_noOutlier))]
head(tpm_filt_counts_noOutlier)
dim(tpm_filt_counts_noOutlier)
#[1] 18639    51
write.csv(tpm_filt_counts_noOutlier, file =paste0(deseq_outDir, "tpm_filt_counts_noOutlier.csv"))


# DESeq design without time-course because two timepoints are very similar
dds1 = DESeqDataSetFromMatrix(countData = tpm_filt_counts_noOutlier,
                             colData = inf_noOutlier,
                             design = design2)

design(dds1)
#~sex + batch + day + condition

dds1 = estimateSizeFactors(dds1)
summary(sizeFactors(dds1))
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#  0.6986  0.9402  1.0032  1.0177  1.1037  1.4019 

# new deseq_outDir since removed outliers
#deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_noOutlier_MW20250106/",sep="")

#saveRDS(dds, file="/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/deseq_results/dds_v2.rds")
#readRDS(dds, file="/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/deseq_results/dds_v2.rds")
saveRDS(dds1, file =paste0(deseq_outDir,"/deseq_results/dds_noOutlier_v1.rds"))
#readRDS(dds1, file =paste0(deseq_outDir,"/deseq_results/dds_noOutlier_v1.rds"))

#plotDispEsts(dds) # doesn't work after adding age_at_sampling to the model, not sure why

vsd = varianceStabilizingTransformation(dds1)

# Remove batch effect only for visualization purpose.
# Important Note: DESeq2 takes raw count numbers as input, and
# models batch effect. Thus, DE gene lists are not affected
# by this removeBatchEffect step.
# See what Mike Love says here: https://support.bioconductor.org/p/104290/
# assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$batch)
#plotPCA(vsd, "batch")

# this is for removng batch effects and covariate effect
design <- model.matrix(~ sex + batch + age_at_sampling + day + condition, data = as.data.frame(colData(vsd)))
assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$batch, design = design)


# Extract a matrix of normalized counts from the S4 object vsd
vsdMat = assay(vsd)

pdf(paste0(deseq_outDir, "/SDplots/tpm.pdf"), width = 7, height = 8)
meanSdPlot(as.matrix(tpm))
dev.off()

pdf(paste0(deseq_outDir, "/SDplots/log_tpm.pdf"), width = 7, height = 8)
meanSdPlot(t(log_tpm))
dev.off()

pdf(paste0(deseq_outDir, "/SDplots/vsdMat.pdf"), width = 7, height = 8)
meanSdPlot(vsdMat)
dev.off()

# PCA using the 500 most variable genes from the variance-stabilized data
#nudge <- position_nudge(x = 2, y = 2)
desired_order_genotype <- c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13")
desired_order_condition <- c("Control", "Within_degron", "Outside_degron", "Truncating")
desired_order_ageAtSampling <- c("kid", "teenager", "adult")

vsd$genotype <- factor(vsd$genotype, levels = desired_order_genotype)
vsd$condition <- factor(vsd$condition, levels = desired_order_condition)
vsd$ageAtSampleing <- factor(vsd$condition, levels = desired_order_ageAtSampling )

#colour_genotype <- c("purple", "green", "blue", "orange")
#names(colour_genotype) <- levels(vsd$genotype)

# no labels
p1 = plotPCA(vsd, intgroup = "condition", ntop = 500) +
  ggplot2::theme(legend.position = "top") + 
  #geom_text_repel(aes(label = name), size = 2, max.overlaps = 50) +
  geom_point(size=0.1)


p2 = plotPCA(vsd, intgroup = "batch", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p3 = plotPCA(vsd, intgroup = "sex", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
    #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p4 = plotPCA(vsd, intgroup = "age_at_sampling", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p5 = plotPCA(vsd, intgroup = "day", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p6 = plotPCA(vsd, intgroup = "genotype", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1) #+ 
  #scale_colour_manual(values = colour_genotype)

pdf(paste0(deseq_outDir, "PCA/PCA_6var_noOutlier_noLabel.pdf"), width = 14, height = 16)
plot_grid(p1, p4, p3, p2, p5, p6, nrow = 3, ncol = 2)
dev.off()

# with labels
p1 = plotPCA(vsd, intgroup = "condition", ntop = 500) +
  ggplot2::theme(legend.position = "top") + 
  geom_text_repel(aes(label = name), size = 2, max.overlaps = 50) +
  geom_point(size=0.1)


p2 = plotPCA(vsd, intgroup = "batch", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p3 = plotPCA(vsd, intgroup = "sex", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p4 = plotPCA(vsd, intgroup = "age_at_sampling", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p5 = plotPCA(vsd, intgroup = "day", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p6 = plotPCA(vsd, intgroup = "genotype", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1) #+ 
  #scale_colour_manual(values = colour_genotype)

pdf(paste0(deseq_outDir, "PCA/PCA_6var_noOutlier_labels.pdf"), width = 20, height = 30)
plot_grid(p1, p4, p3, p2, p5, p6, nrow = 3, ncol = 2)
dev.off()


# plot_grid(p1, p4, p3, p2, nrow = 2, ncol = 2)
# ggsave(paste0(deseq_outDir, "/PCA/PCA_4categories_v3_label.pdf"), width = 15, height = 10)

# plot umap to see the relationships better
# plot with the vsdMat matrix (contains the variance-stabilized, batch-effect-corrected normalized counts.)
# transpose data since umap needssamples as rows and features (genes) as columns
umap_input <- t(vsdMat)

# Run UMAP
set.seed(42)  # For reproducibility
umap_results <- umap(umap_input)

# Convert UMAP results to a data frame
umap_df <- as.data.frame(umap_results$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

# Add sample metadata (from `inf_noOutlier`)
umap_df$sample <- rownames(umap_df)
umap_df <- cbind(umap_df, inf_noOutlier)

# plot umap with labels
a1=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = day)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Day"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a2=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = genotype)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Genotype"    # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 3, 4, 8, 6),  # Shapes for each genotype
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Y_fs*13")
  ) +
  theme(legend.position = "right")

a3=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = genotype, shape = condition)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Genotype",  # Custom label for the color legend
    shape = "Condition"  # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("WT" = "azure4", 
               "D868N" = "orange", 
               "I871T" = "darkorange4", 
               "E858K" = "deepskyblue", 
               "T962del" = "chartreuse2", 
               "K693P_fs*86" = "darkorchid1", 
               "I822Y_fs*13" = "darkorchid4"),
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Yfs*13")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 8),  # Shapes for each condition
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")



a4=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = sex)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Sex"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a5=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Batch"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a6=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = age_at_sampling)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Age at sampling"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15),  # Shapes for each condition
    breaks = c("kid", "teenager", "adult"),
    labels = c("Kid", "Teenager", "Adult")
  ) +
  theme(legend.position = "right")

pdf(paste0(deseq_outDir, "UMAP/UMAP_6var_noOutlier_labels.pdf"), width = 20, height = 20)
plot_grid(a1,a2,a3,a4,a5,a6, nrow = 3, ncol = 2)
dev.off()

# plot umap without labels
a1=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = day)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Day"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a2=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = genotype)) +
  geom_point(size = 3) +
 # geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Genotype"    # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 3, 4, 8, 6),  # Shapes for each genotype
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Y_fs*13")
  ) +
  theme(legend.position = "right")

a3=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = genotype, shape = condition)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Genotype",  # Custom label for the color legend
    shape = "Condition"  # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("WT" = "azure4", 
               "D868N" = "orange", 
               "I871T" = "darkorange4", 
               "E858K" = "deepskyblue", 
               "T962del" = "chartreuse2", 
               "K693P_fs*86" = "darkorchid1", 
               "I822Y_fs*13" = "darkorchid4"),
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Yfs*13")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 8),  # Shapes for each condition
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")



a4=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = sex)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Sex"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a5=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Batch"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a6=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = age_at_sampling)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Age at sampling"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15),  # Shapes for each condition
    breaks = c("kid", "teenager", "adult"),
    labels = c("Kid", "Teenager", "Adult")
  ) +
  theme(legend.position = "right")

pdf(paste0(deseq_outDir, "UMAP/UMAP_6var_noOutlier_nolabel.pdf"), width = 10, height = 10)
plot_grid(a1,a2,a3,a4,a5,a6, nrow = 3, ncol = 2)
dev.off()

```
look okay after removing outliers
continue to remove d0 then do deseq2 comparing condition
```{r DEanalysis-part1-3}
# new deseq_outDir since removed outliers and d0
deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_noOutlier_noD0_MW20250107/1_condition/",sep="")

head(inf)

# # Make sure that the IDs in the info table have
# # the same order as the columns in the count table
# rownames(inf) = inf$ID # make ID rownames
# colnames(rcounts) = str_replace(colnames(rcounts), ".bam", "") # remove ".bam" from colnames
# colnames(tpm_filt_counts) = str_replace(colnames(tpm_filt_counts), ".bam", "") # remove ".bam" from colnames
# tpm_filt_counts = tpm_filt_counts[, rownames(inf)] # order columns based on sample list in inf
# 
# colnames(tpm_filt_counts) == rownames(inf) # check if it worked (all should be TRUE, yes)
# colnames(tpm_filt_counts) = inf$ID # make IDs in info as colnames of filtered counts, data still the same
# rownames(inf) = inf$ID # make IDs back to rowname
# colnames(tpm_filt_counts) == rownames(inf) # check if it worked (all should be TRUE, yes, hope everything is fine)
# 
# nrow(tpm_filt_counts)
# #[1] 18639
# 
# inf$sex <- factor(inf$sex)
# inf$batch <- factor(inf$batch)
# inf$day <- factor(inf$day)
# inf$condition <- factor(inf$condition)
# inf$genotype <- factor(inf$genotype)

# designs
design1 = ~ sex + batch + day
design2= ~ sex + batch + day + condition
design3 = ~ sex + batch + day + condition + condition:day

# # remove samples that are outlier
# exclude_ids <- c("B8D10_1223", "B7D12_1225", "B7D12_1223")
# inf_noOutlier <- inf[!inf$ID %in% exclude_ids, ]
# #51 samples left

# remove d0 for the rest of deseq2
inf_noOutlier_noD0 <- inf_noOutlier[inf_noOutlier$day != "d0", ]
# 45 samples left

# remove d0 from the tpm_filt_counts_noOutlier
# Subset tpm_filt_counts to match the IDs in inf_noOutlier_noD0
tpm_filt_counts_noOutlier_noD0 <- tpm_filt_counts[, colnames(tpm_filt_counts) %in% inf_noOutlier_noD0$ID]

# Ensure the column order of tpm_filt_counts matches inf_subset
tpm_filt_counts_noOutlier_noD0 <- tpm_filt_counts_noOutlier_noD0[, match(inf_noOutlier_noD0$ID, colnames(tpm_filt_counts_noOutlier_noD0))]
head(tpm_filt_counts_noOutlier_noD0)
dim(tpm_filt_counts_noOutlier_noD0)
#[1] 18639    45
write.csv(tpm_filt_counts_noOutlier_noD0, file =paste0(deseq_outDir, "tpm_filt_counts_noOutlier_noD0.csv"))


# DESeq design without time-course because two timepoints are very similar
dds3 = DESeqDataSetFromMatrix(countData = tpm_filt_counts_noOutlier_noD0,
                             colData = inf_noOutlier_noD0,
                             design = design2)

design(dds3)
#~sex + batch + day + condition

dds3 = estimateSizeFactors(dds3)
summary(sizeFactors(dds3))
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#  0.7145  0.9245  0.9997  1.0159  1.0934  1.4351 


#saveRDS(dds, file="/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/deseq_results/dds_v2.rds")
#readRDS(dds, file="/data/workspaces/lag/workspaces/lg-scrna/working_data/gokberk/setbp1/results/DESeq/deseq_results/dds_v2.rds")
#saveRDS(dds1, file =paste0(deseq_outDir,"/deseq_results/dds_noOutlier_v1.rds"))
#readRDS(dds1, file =paste0(deseq_outDir,"/deseq_results/dds_noOutlier_v1.rds"))
saveRDS(dds3, file =paste0(deseq_outDir,"/deseq_results/dds_noOutlier_noD0_v1.rds"))
#readRDS(dds3, file =paste0(deseq_outDir,"/deseq_results/dds_noOutlier_noD0_v1.rds"))

#plotDispEsts(dds) # doesn't work after adding age_at_sampling to the model, not sure why

vsd = varianceStabilizingTransformation(dds3)

# Remove batch effect only for visualization purpose.
# Important Note: DESeq2 takes raw count numbers as input, and
# models batch effect. Thus, DE gene lists are not affected
# by this removeBatchEffect step.
# See what Mike Love says here: https://support.bioconductor.org/p/104290/
# assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$batch)
# plotPCA(vsd, "batch")

# this is for removng batch effects and covariate effect
design <- model.matrix(~ sex + batch + age_at_sampling + day + condition, data = as.data.frame(colData(vsd)))
assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$batch, design = design)
# doesnt seem to differ a lot with different covariates so there is likely something that we don't know about

# Extract a matrix of normalized counts from the S4 object vsd
vsdMat = assay(vsd)

pdf(paste0(deseq_outDir, "/SDplots/tpm.pdf"), width = 7, height = 8)
meanSdPlot(as.matrix(tpm))
dev.off()

pdf(paste0(deseq_outDir, "/SDplots/log_tpm.pdf"), width = 7, height = 8)
meanSdPlot(t(log_tpm))
dev.off()

pdf(paste0(deseq_outDir, "/SDplots/vsdMat.pdf"), width = 7, height = 8)
meanSdPlot(vsdMat)
dev.off()

# PCA using the 500 most variable genes from the variance-stabilized data
#nudge <- position_nudge(x = 2, y = 2)
desired_order_genotype <- c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13")
desired_order_condition <- c("Control", "Within_degron", "Outside_degron", "Truncating")
desired_order_ageAtSampling <- c("kid", "teenager", "adult")

vsd$genotype <- factor(vsd$genotype, levels = desired_order_genotype)
vsd$condition <- factor(vsd$condition, levels = desired_order_condition)
vsd$ageAtSampleing <- factor(vsd$condition, levels = desired_order_ageAtSampling )

#colour_genotype <- c("purple", "green", "blue", "orange")
#names(colour_genotype) <- levels(vsd$genotype)

# no labels
p1 = plotPCA(vsd, intgroup = "condition", ntop = 500) +
  ggplot2::theme(legend.position = "top") + 
  #geom_text_repel(aes(label = name), size = 2, max.overlaps = 50) +
  geom_point(size=0.1)


p2 = plotPCA(vsd, intgroup = "batch", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p3 = plotPCA(vsd, intgroup = "sex", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
    #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p4 = plotPCA(vsd, intgroup = "age_at_sampling", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p5 = plotPCA(vsd, intgroup = "day", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p6 = plotPCA(vsd, intgroup = "genotype", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  #geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1) #+ 
  #scale_colour_manual(values = colour_genotype)

pdf(paste0(deseq_outDir, "PCA/PCA_6var_noOutlierNoD0_noLabel.pdf"), width = 14, height = 16)
plot_grid(p1, p4, p3, p2, p5, p6, nrow = 3, ncol = 2)
dev.off()

# with labels
p1 = plotPCA(vsd, intgroup = "condition", ntop = 500) +
  ggplot2::theme(legend.position = "top") + 
  geom_text_repel(aes(label = name), size = 2, max.overlaps = 50) +
  geom_point(size=0.1)


p2 = plotPCA(vsd, intgroup = "batch", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p3 = plotPCA(vsd, intgroup = "sex", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p4 = plotPCA(vsd, intgroup = "age_at_sampling", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p5 = plotPCA(vsd, intgroup = "day", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1)

p6 = plotPCA(vsd, intgroup = "genotype", ntop = 500) +
  ggplot2::theme(legend.position = "top") +
  geom_text_repel(aes(label = name), size = 2) +
  geom_point(size=0.1) #+ 
  #scale_colour_manual(values = colour_genotype)

pdf(paste0(deseq_outDir, "PCA/PCA_6var_noOutlierNoD0_labels.pdf"), width = 20, height = 30)
plot_grid(p1, p4, p3, p2, p5, p6, nrow = 3, ncol = 2)
dev.off()


# plot_grid(p1, p4, p3, p2, nrow = 2, ncol = 2)
# ggsave(paste0(deseq_outDir, "/PCA/PCA_4categories_v3_label.pdf"), width = 15, height = 10)

# plot umap to see the relationships better
# plot with the vsdMat matrix (contains the variance-stabilized, batch-effect-corrected normalized counts.)
# transpose data since umap needssamples as rows and features (genes) as columns
umap_input <- t(vsdMat)

# Run UMAP
set.seed(42)  # For reproducibility
umap_results <- umap(umap_input)

# Convert UMAP results to a data frame
umap_df <- as.data.frame(umap_results$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

# Add sample metadata (from `inf_noOutlier_noD0`)
umap_df$sample <- rownames(umap_df)
umap_df <- cbind(umap_df, inf_noOutlier_noD0)

# plot umap with labels
a1=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = day)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Day"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a2=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = genotype)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Genotype"    # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 3, 4, 8, 6),  # Shapes for each genotype
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Y_fs*13")
  ) +
  theme(legend.position = "right")

a3=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = genotype, shape = condition)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Genotype",  # Custom label for the color legend
    shape = "Condition"  # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("WT" = "azure4", 
               "D868N" = "orange", 
               "I871T" = "darkorange4", 
               "E858K" = "deepskyblue", 
               "T962del" = "chartreuse2", 
               "K693P_fs*86" = "darkorchid1", 
               "I822Y_fs*13" = "darkorchid4"),
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Yfs*13")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 8),  # Shapes for each condition
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")



a4=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = sex)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Sex"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a5=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Batch"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a6=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = age_at_sampling)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Age at sampling"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15),  # Shapes for each condition
    breaks = c("kid", "teenager", "adult"),
    labels = c("Kid", "Teenager", "Adult")
  ) +
  theme(legend.position = "right")

pdf(paste0(deseq_outDir, "UMAP/UMAP_6var_noOutlierNoD0_labels.pdf"), width = 20, height = 20)
plot_grid(a1,a2,a3,a4,a5,a6, nrow = 3, ncol = 2)
dev.off()

# plot umap without labels
a1=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = day)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Day"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a2=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = genotype)) +
  geom_point(size = 3) +
 # geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Genotype"    # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 3, 4, 8, 6),  # Shapes for each genotype
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Y_fs*13")
  ) +
  theme(legend.position = "right")

a3=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = genotype, shape = condition)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Genotype",  # Custom label for the color legend
    shape = "Condition"  # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("WT" = "azure4", 
               "D868N" = "orange", 
               "I871T" = "darkorange4", 
               "E858K" = "deepskyblue", 
               "T962del" = "chartreuse2", 
               "K693P_fs*86" = "darkorchid1", 
               "I822Y_fs*13" = "darkorchid4"),
    breaks = c("WT", "D868N", "I871T", "E858K", "T962del", "K693P_fs*86", "I822Y_fs*13"),
    labels = c("WT", "D868N", "I871T", "E858K", "T962del", "K693Pfs*86", "I822Yfs*13")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 8),  # Shapes for each condition
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")



a4=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = sex)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Sex"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a5=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Batch"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  theme(legend.position = "right")

a6=ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = condition, shape = age_at_sampling)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = ID), size = 2, max.overlaps = 100) +  # Add sample names
  #theme_minimal() +
  theme_grey() +
  labs(
    title = "UMAP of Variance-Stabilized Counts",
    x = "UMAP Dimension 1",
    y = "UMAP Dimension 2",
    color = "Condition",  # Custom label for the color legend
    shape = "Age at sampling"         # Custom label for the shape legend
  ) +
  scale_color_manual(
    values = c("Control" = "azure4", "Within_degron" = "orange", 
               "Outside_degron" = "deepskyblue", "Truncating" = "darkorchid1"),
    breaks = c("Control", "Within_degron", "Outside_degron", "Truncating"),
    labels = c("Control", "Within degron", "Outside degron", "Truncating")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15),  # Shapes for each condition
    breaks = c("kid", "teenager", "adult"),
    labels = c("Kid", "Teenager", "Adult")
  ) +
  theme(legend.position = "right")

pdf(paste0(deseq_outDir, "UMAP/UMAP_6var_noOutlierNoD0_nolabel.pdf"), width = 10, height = 10)
plot_grid(a1,a2,a3,a4,a5,a6, nrow = 3, ncol = 2)
dev.off()

```
umap Show better relationship between samples
include d0  control lines versus d10/12  control lines
continue  with dss3

```{r DEanalysis-part2}

deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/4_deseq2_noOutlier_noD0_MW20250107/1_condition/",sep="")

# Run DESeq for different combinations of conditions
# to make volcano plots
# continue with dds3 for only iNs
#dds2 = DESeq(dds, betaPrior = FALSE)
dds2 = DESeq(dds3, betaPrior = FALSE)
saveRDS(dds2, file =paste0(deseq_outDir,"deseq_results/dds2_d10-12iN_only.rds"))
#readRDS(dds2, file =paste0(deseq_outDir,"deseq_results/dds2_d10-12iN_only.rds"))

# Control vs. Outside_degron
res1 = results(dds2, contrast = c("condition", "Outside_degron", "Control"))
res1 = lfcShrink(dds2, contrast = c("condition", "Outside_degron", "Control"), res=res1, type = "normal")
res1$symbol = mapIds(org.Hs.eg.db, keys=row.names(res1), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
write.csv(as.data.frame(res1), paste0(deseq_outDir,
                                     "/deseq_results/res_ControlvsOutsideDegron.csv"), quote = F)
#removeNA
res1_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_ControlvsOutsideDegron.csv"), header = T)
res1_deseq_results_noNA = na.omit(res1_deseq_results)
nrow(res1_deseq_results_noNA)
write.csv(as.data.frame(res1_deseq_results_noNA), paste0(deseq_outDir,
                                     "/deseq_results/res_ControlvsOutsideDegron_noNA.csv"), quote = F)
saveRDS(res1, file =paste0(deseq_outDir,"deseq_results/res1_condition_ControlvsOutsideDegron.rds"))


# Control vs. Within_degron
res2 = results(dds2, contrast = c("condition", "Within_degron", "Control"))
res2 = lfcShrink(dds2, contrast = c("condition", "Within_degron", "Control"), res=res2, type = "normal")
res2$symbol = mapIds(org.Hs.eg.db, keys=row.names(res2), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
write.csv(as.data.frame(res2), paste0(deseq_outDir,
                                     "/deseq_results/res_ControlvsWithin_degronSGS.csv"), quote = F)

#removeNA
res2_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_ControlvsWithin_degronSGS.csv"), header = T)
res2_deseq_results_noNA = na.omit(res2_deseq_results)
nrow(res2_deseq_results_noNA)
write.csv(as.data.frame(res2_deseq_results_noNA), paste0(deseq_outDir,
                                     "/deseq_results/res_ControlvsWithin_degronSGS_noNA.csv"), quote = F)
saveRDS(res2, file =paste0(deseq_outDir,"deseq_results/res2_condition_ControlvsWithin_degronSGS.rds"))


# Control vs. Truncating
res3 = results(dds2, contrast = c("condition", "Truncating", "Control"))
res3 = lfcShrink(dds2, contrast = c("condition", "Truncating", "Control"), res=res3, type = "normal")
res3$symbol = mapIds(org.Hs.eg.db, keys=row.names(res3), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
write.csv(as.data.frame(res3), paste0(deseq_outDir,
                                     "/deseq_results/res_ControlvsTruncating.csv"), quote = F)

#removeNA
res3_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_ControlvsTruncating.csv"), header = T)
res3_deseq_results_noNA = na.omit(res3_deseq_results)
nrow(res3_deseq_results_noNA)
write.csv(as.data.frame(res3_deseq_results_noNA), paste0(deseq_outDir,
                                     "/deseq_results/res_ControlvsTruncating_noNA.csv"), quote = F)
saveRDS(res3, file =paste0(deseq_outDir,"deseq_results/res3_ControlvsTruncating.rds"))


# # Control vs. Outside
# res4 = results(dds2, contrast = c("condition", "Control", "Outside"))
# res4 = lfcShrink(dds2, contrast = c("condition", "Control", "Outside"), res=res4, type = "normal")
# res4$symbol = mapIds(org.Hs.eg.db, keys=row.names(res4), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
# write.csv(as.data.frame(res4), paste0(deseq_outDir,
#                                     "/deseq_results/res_ControlvsOutside.csv"), quote = F)

# Outside_degron vs. Within_degron
res5 = results(dds2, contrast = c("condition", "Outside_degron", "Within_degron"))
res5 = lfcShrink(dds2, contrast = c("condition", "Outside_degron", "Within_degron"), res=res5, type = "normal")
res5$symbol = mapIds(org.Hs.eg.db, keys=row.names(res5), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
write.csv(as.data.frame(res5), paste0(deseq_outDir,
                                     "/deseq_results/res_Outside_degronvsWithin_degronSGS.csv"), quote = F)

#removeNA
res5_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_Outside_degronvsWithin_degronSGS.csv"), header = T)
res5_deseq_results_noNA = na.omit(res5_deseq_results)
nrow(res5_deseq_results_noNA)
write.csv(as.data.frame(res5_deseq_results_noNA), paste0(deseq_outDir,
                                     "/deseq_results/res_Outside_degronvsWithin_degronSGS_noNA.csv"), quote = F)


# Within_degron vs. truncating
res6 = results(dds2, contrast = c("condition", "Within_degron", "Truncating"))
res6 = lfcShrink(dds2, contrast = c("condition", "Within_degron", "Truncating"), res=res6, type = "normal")
res6$symbol = mapIds(org.Hs.eg.db, keys=row.names(res6), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
write.csv(as.data.frame(res6), paste0(deseq_outDir,
                                     "/deseq_results/res_within_degronSGSvstruncating.csv"), quote = F)

#removeNA
res6_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_within_degronSGSvstruncating.csv"), header = T)
res6_deseq_results_noNA = na.omit(res6_deseq_results)
nrow(res6_deseq_results_noNA)
write.csv(as.data.frame(res6_deseq_results_noNA), paste0(deseq_outDir,
                                     "/deseq_results/res_within_degronSGSvstruncating_noNA.csv"), quote = F)

# Outside_degron vs. truncating
res7 = results(dds2, contrast = c("condition", "Outside_degron", "Truncating"))
res7 = lfcShrink(dds2, contrast = c("condition", "Outside_degron", "Truncating"), res=res7, type = "normal")
res7$symbol = mapIds(org.Hs.eg.db, keys=row.names(res7), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
write.csv(as.data.frame(res7), paste0(deseq_outDir,
                                     "/deseq_results/res_Outside_degronvstruncating.csv"), quote = F)

#removeNA
res7_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_Outside_degronvstruncating.csv"), header = T)
res7_deseq_results_noNA = na.omit(res7_deseq_results)
nrow(res7_deseq_results_noNA)
write.csv(as.data.frame(res7_deseq_results_noNA), paste0(deseq_outDir,
                                     "/deseq_results/res_Outside_degronvstruncating_noNA.csv"), quote = F)

# [1] 14481 in noNA
###################################################

# # these are for when including d0 samples in another script
# # d0 control vs d10/12 control
# res8 = results(dds2, contrast = c("ID3", "control_d10/12", "control_d0"))
# res8 = lfcShrink(dds2, contrast = c("ID3", "control_d10/12", "control_d0"), res=res8, type = "normal")
# res8$symbol = mapIds(org.Hs.eg.db, keys=row.names(res8), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
# write.csv(as.data.frame(res8), paste0(deseq_outDir,
#                                      "/deseq_results/res_d0controlvsd10-12control.csv"), quote = F)
# 
# #removeNA
# res8_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_d0controlvsd10-12control.csv"), header = T)
# res8_deseq_results_noNA = na.omit(res8_deseq_results)
# nrow(res8_deseq_results_noNA)
# #[1]  11819
# write.csv(as.data.frame(res8_deseq_results_noNA), paste0(deseq_outDir,
#                                      "/deseq_results/res_d0controlvsd10-12control_noNA.csv"), quote = F)
# 
# # d0 vs d10 controls
# res9 = results(dds2, contrast = c("ID2", "control_d10", "control_d0"))
# res9 = lfcShrink(dds2, contrast = c("ID2", "control_d10", "control_d0"), res=res9, type = "normal")
# res9$symbol = mapIds(org.Hs.eg.db, keys=row.names(res9), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
# write.csv(as.data.frame(res9), paste0(deseq_outDir,
#                                      "/deseq_results/res_d0controlvsd10control.csv"), quote = F)
# 
# #removeNA
# res9_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_d0controlvsd10control.csv"), header = T)
# res9_deseq_results_noNA = na.omit(res9_deseq_results)
# nrow(res9_deseq_results_noNA)
# #[1]  11819
# write.csv(as.data.frame(res9_deseq_results_noNA), paste0(deseq_outDir,
#                                      "/deseq_results/res_d0controlvsd10control_noNA.csv"), quote = F)
# 
# # d0 vs d12 controls
# res10 = results(dds2, contrast = c("ID2", "control_d12", "control_d0"))
# res10 = lfcShrink(dds2, contrast = c("ID2", "control_d12", "control_d0"), res=res10, type = "normal")
# res10$symbol = mapIds(org.Hs.eg.db, keys=row.names(res10), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
# write.csv(as.data.frame(res10), paste0(deseq_outDir,
#                                      "/deseq_results/res_d0controlvsd12control.csv"), quote = F)
# 
# #removeNA
# res10_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_d0controlvsd12control.csv"), header = T)
# res10_deseq_results_noNA = na.omit(res10_deseq_results)
# nrow(res10_deseq_results_noNA)
# #[1]  11819
# write.csv(as.data.frame(res10_deseq_results_noNA), paste0(deseq_outDir,
#                                      "/deseq_results/res_d0controlvsd12control_noNA.csv"), quote = F)
# 
# # d0 vs d12 controls
# res11 = results(dds2, contrast = c("ID2", "control_d12", "control_d10"))
# res11 = lfcShrink(dds2, contrast = c("ID2", "control_d12", "control_d10"), res=res11, type = "normal")
# res11$symbol = mapIds(org.Hs.eg.db, keys=row.names(res11), column="SYMBOL", keytype="ENSEMBL", multiVals="first")
# write.csv(as.data.frame(res11), paste0(deseq_outDir,
#                                      "/deseq_results/res_d10controlvsd12control.csv"), quote = F)
# 
# #removeNA
# res11_deseq_results = read.csv(paste0(deseq_outDir, "/deseq_results/res_d10controlvsd12control.csv"), header = T)
# res11_deseq_results_noNA = na.omit(res11_deseq_results)
# nrow(res11_deseq_results_noNA)
# #[1]  11819
# write.csv(as.data.frame(res11_deseq_results_noNA), paste0(deseq_outDir,
#                                      "/deseq_results/res_d10controlvsd12control_noNA.csv"), quote = F)

```


```{r volcanoPlots}

# Plot and save volcanoplots

pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-ControlvsOutsideDegron.pdf"), width = 7, height = 8)
EnhancedVolcano(res1, lab = NA, x="log2FoldChange", y="padj",
                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
                gridlines.minor = FALSE, gridlines.major = FALSE)
dev.off()

pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-ControlvsWithin_degronSGS.pdf"), width = 7, height = 8)
EnhancedVolcano(res2, lab = NA, x="log2FoldChange", y="padj",
                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
                gridlines.minor = FALSE, gridlines.major = FALSE)
dev.off()

pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-ControlvsTruncating.pdf"), width = 7, height = 8)
EnhancedVolcano(res3, lab = NA, x="log2FoldChange", y="padj",
                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
                gridlines.minor = FALSE, gridlines.major = FALSE)
dev.off()

#pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-ControlvsOutside.pdf"), width = 7, height = 8)
#EnhancedVolcano(res4, lab = NA, x="log2FoldChange", y="padj",
#                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
#                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
#                gridlines.minor = FALSE, gridlines.major = FALSE, )
#dev.off()

pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-Outside_degronvsWithin_degronSGS.pdf"), width = 7, height = 8)
EnhancedVolcano(res5, lab = NA, x="log2FoldChange", y="padj",
                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
                gridlines.minor = FALSE, gridlines.major = FALSE)
dev.off()

pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-within_degronSGSvstruncating.pdf"), width = 7, height = 8)
EnhancedVolcano(res6, lab = NA, x="log2FoldChange", y="padj",
                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
                gridlines.minor = FALSE, gridlines.major = FALSE)
dev.off()

pdf(paste0(deseq_outDir, "volcanoPlots/Volcano-Outside_degronvstruncating.pdf"), width = 7, height = 8)
EnhancedVolcano(res7, lab = NA, x="log2FoldChange", y="padj",
                xlim = c(-2, 2), ylim = c(0,15), captionLabSize = 9,
                col=c('grey75', 'grey50', 'grey25', 'red','blue'), pCutoff = 0.1, FCcutoff = 0.6,
                gridlines.minor = FALSE, gridlines.major = FALSE)
dev.off()


```

```{r getDEgeneCounts}

get_upregulated = function(df){

    df = as.data.frame(df)
  
    key = intersect(rownames(df)[which(df$log2FoldChange>=1)], rownames(df)[which(df$padj<=0.05)])

    results = as.data.frame((df)[which(rownames(df) %in% key),])
    return(results)
}

get_downregulated = function(df){

    df = as.data.frame(df)
  
    key = intersect(rownames(df)[which(df$log2FoldChange<=-1)], rownames(df)[which(df$padj<=0.05)])

    results = as.data.frame((df)[which(rownames(df) %in% key),])
    return(results)
}

for (i in 1:7) {
  if (i==4) next # skip ControlvsOutside
  up_list = get_upregulated(get(paste0("res", i)))
  write.table(up_list, paste0(deseq_outDir,"/sigDEgeneLists/sigUpRegulatedGenes_res", i, ".tab"),
              quote = F, sep = "\t")
  
  }

for (i in 1:7) {
  if (i==4) next # skip ControlvsOutside
  down_list = get_downregulated(get(paste0("res", i)))
  write.table(down_list, paste0(deseq_outDir,"/sigDEgeneLists/sigDownRegulatedGenes_res", i, ".tab"),
              quote = F, sep = "\t")
  
  }
#check number of up- and down- regulated genes for each comparison

for (i in 1:7) {
  if (i==4) next # skip ControlvsOutside
  summary_list = summary(get(paste0("res", i)))}

  #write.table(summary_list, paste0(deseq_outDir,"/sigDEgeneLists/summary_res", i, ".tab"),
 #             quote = F, sep = "\t")} #don't know why can't put in a table

outside_in_up = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigUpRegulatedGenes_res5.tab"), header = T)
outside_in_down = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigDownRegulatedGenes_res5.tab"), header = T)

inside_truncating_up = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigUpRegulatedGenes_res6.tab"), header = T)
inside_truncating_down = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigDownRegulatedGenes_res6.tab"), header = T)

outside_truncating_up = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigUpRegulatedGenes_res7.tab"), header = T)
outside_truncating_down = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigDownRegulatedGenes_res7.tab"), header = T)

################################################################
# get all degs, sigup & sigdown genes (MW)
get_all_degs = function(df) {
    df = as.data.frame(df)
    
    # Remove rows with NA in log2FoldChange, padj, or symbol
    df = df[!is.na(df$log2FoldChange) & !is.na(df$padj) & !is.na(df$symbol), ]
    
    # Get upregulated genes
    up_key = intersect(rownames(df)[which(df$log2FoldChange >= 1)], rownames(df)[which(df$padj <= 0.05)])
    up_results = as.data.frame(df[which(rownames(df) %in% up_key),])
    
    # Get downregulated genes
    down_key = intersect(rownames(df)[which(df$log2FoldChange <= -1)], rownames(df)[which(df$padj <= 0.05)])
    down_results = as.data.frame(df[which(rownames(df) %in% down_key),])
    
    # Combine results
    all_results = rbind(up_results, down_results)
    return(all_results)
}

# Loop to save results for all res objects as csv
for (i in 1:7) {
    if (i==4) next # skip ControlvsOutside

    res_obj = get(paste0("res", i))
    
    # Get all DEGs
    all_degs = get_all_degs(res_obj)
    
    # Write combined DEGs to CSV file
    write.csv(all_degs, file = paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_sigAllDEGenes.csv"),
              row.names = TRUE)
    
    # Optionally, save upregulated and downregulated separately
    up_list = get_upregulated(res_obj)
    write.csv(up_list, file = paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_sigUpRegulatedGenes.csv"),
              row.names = TRUE)
    
    down_list = get_downregulated(res_obj)
    write.csv(down_list, file = paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_sigDownRegulatedGenes.csv"),
              row.names = TRUE)
}

# Loop to save results for all res objects as tab
for (i in 1:7) {
    if (i==4) next # skip ControlvsOutside

    res_obj = get(paste0("res", i))
    
    # Get all DEGs
    all_degs = get_all_degs(res_obj)
    
    # Write combined DEGs to file
    write.table(all_degs, paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_sigAllDEGenes.tab"),
                quote = FALSE, sep = "\t")
    
    # Optionally, save upregulated and downregulated separately
    up_list = get_upregulated(res_obj)
    write.table(up_list, paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_sigUpRegulatedGenes.tab"),
                quote = FALSE, sep = "\t")
    
    down_list = get_downregulated(res_obj)
    write.table(down_list, paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_sigDownRegulatedGenes.tab"),
                quote = FALSE, sep = "\t")
}

#check number of up- and down- regulated genes for each comparison
# Loop for summary statistics
for (i in 1:7) {
    if (i==4) next # skip ControlvsOutside

  # Capture the output of summary() to a character vector
  summary_output <- capture.output(summary(get(paste0("res", i))))
  
  # Write the captured output to a file
  writeLines(summary_output, paste0(deseq_outDir, "sigDEgeneListsMW/res", i, "_summary.txt"))
}

```
DESeq2 applies independent filtering during the results() step, based on the mean normalized counts of genes.
This filtering ensures that low-information genes (e.g., genes with very low counts) are excluded from multiple testing correction.

so no need to re-do filtering

```{r heatmaps}

# We can’t plot the entire data set, let’s just select the top 300 by padj
# Get top 300 differentially expressed genes between Control and Outside_degron samples

# First, add gene IDs as a column
res1$geneIDs = rownames(res1)

sigGenes_ControlvsOutside_degron = as.data.frame(res1) %>% 
                                   top_n(300, wt=-padj) %>% 
                                   pull("geneIDs")

# We’ll use normalised expression values, so we’ll use the vst function -> Changed this, see below
#plotDat_ControlvsOutside_degron <- vst(dds)[sigGenes_ControlvsOutside_degron,] %>% 
#  assay()
# Here we used Batch Effect Removed (with limma) vsd file to see heatmaps without batch effect
plotDat_ControlvsOutside_degron <- vsd[sigGenes_ControlvsOutside_degron,] %>% 
  assay()

z.mat_ControlvsOutside_degron <- t(scale(t(plotDat_ControlvsOutside_degron), center=TRUE, scale=TRUE))

myPalette <- c("royalblue3", "ivory", "orangered3")
myRamp <- colorRamp2(c(-2, 0, 2), myPalette)

colData(dds3)$condition <- factor(colData(dds3)$condition, levels = c("Control", "Within_degron", "Outside_degron", "Truncating"))

ha = HeatmapAnnotation(df = colData(dds3)[,c("condition", "batch", "sex")],
                        col = list(condition = c("Control" = "azure4",
                                                 "Outside_degron"="deepskyblue", "Truncating" = "darkorchid1",
                                                 "Within_degron" = "orange"),
                                   batch = c("b6" = "lightblue", "b7" = "darkblue", "b8" = "blue"),
                                   sex = c("M" = "orange4", "F" = "darkgreen")))

pdf(file = paste0(deseq_outDir, "/heatmaps/heatmap_top300DEgenes_ControlvsOutside_degron_batchEffRemoved.pdf"),
    width = 12, height = 8)
Heatmap(z.mat_ControlvsOutside_degron, name = "z-score",
        col = myRamp,            
        show_row_name = FALSE,
        #split=3,
        rect_gp = gpar(col = "lightgrey", lwd=0.3),
        top_annotation = ha)
dev.off()

# Get top 300 differentially expressed genes between Control and Within_degron

res2$geneIDs = rownames(res2)

sigGenes_ControlvsWithin_degronSGS = as.data.frame(res2) %>% 
                                   top_n(300, wt=-padj) %>% 
                                   pull("geneIDs")

#plotDat_ControlvsWithin_degronSGS <- vst(dds)[sigGenes_ControlvsWithin_degronSGS,] %>% 
#  assay()
plotDat_ControlvsWithin_degronSGS <- vsd[sigGenes_ControlvsWithin_degronSGS,] %>% 
  assay()
z.mat_ControlvsWithin_degronSGS <- t(scale(t(plotDat_ControlvsWithin_degronSGS), center=TRUE, scale=TRUE))

pdf(file = paste0(deseq_outDir, "/heatmaps/heatmap_top300DEgenes_ControlvsWithin_degronSGS_batchEffRemoved.pdf"),
    width = 12, height = 8)
Heatmap(z.mat_ControlvsWithin_degronSGS, name = "z-score",
        col = myRamp,            
        show_row_name = FALSE,
        #split=3,
        rect_gp = gpar(col = "lightgrey", lwd=0.3),
        top_annotation = ha)
dev.off()

# Get top 300 differentially expressed genes between Control vs. Truncating

res3$geneIDs = rownames(res3)

sigGenes_ControlvsTruncating = as.data.frame(res3) %>% 
                                   top_n(300, wt=-padj) %>% 
                                   pull("geneIDs")

#plotDat_ControlvsTruncating <- vst(dds)[sigGenes_ControlvsTruncating,] %>% 
#  assay()
plotDat_ControlvsTruncating <- vsd[sigGenes_ControlvsTruncating,] %>% 
  assay()
z.mat_ControlvsTruncating <- t(scale(t(plotDat_ControlvsTruncating), center=TRUE, scale=TRUE))

pdf(file = paste0(deseq_outDir, "/heatmaps/heatmap_top300DEgenes_ControlvsTruncating_batchEffRemoved.pdf"),
    width = 12, height = 8)
Heatmap(z.mat_ControlvsTruncating, name = "z-score",
        col = myRamp,            
        show_row_name = FALSE,
        #split=3,
        rect_gp = gpar(col = "lightgrey", lwd=0.3),
        top_annotation = ha)
dev.off()

# Get top 300 differentially expressed genes between Control vs. Outside

#res4$geneIDs = rownames(res4)

#sigGenes_ControlvsOutside = as.data.frame(res4) %>% 
#                                   top_n(300, wt=-padj) %>% 
#                                   pull("geneIDs")

#plotDat_ControlvsOutside <- vst(dds)[sigGenes_ControlvsOutside,] %>% 
#  assay()
#plotDat_ControlvsOutside <- vsd[sigGenes_ControlvsOutside,] %>% 
#  assay()
#z.mat_ControlvsOutside <- t(scale(t(plotDat_ControlvsOutside), center=TRUE, scale=TRUE))

#pdf(file = paste0(deseq_outDir, "heatmap_top300DEgenes_ControlvsOutside_batchEffRemoved.pdf"),
#    width = 12, height = 8)
#Heatmap(z.mat_ControlvsOutside, name = "z-score",
#        col = myRamp,            
#        show_row_name = FALSE,
#        #split=3,
#        rect_gp = gpar(col = "lightgrey", lwd=0.3),
#        top_annotation = ha)
#dev.off()

# Get top 300 differentially expressed genes between Outside_degron vs. Within_degron

res5$geneIDs = rownames(res5)

sigGenes_Outside_degronvsWithin_degronSGS = as.data.frame(res5) %>% 
                                   top_n(300, wt=-padj) %>% 
                                   pull("geneIDs")

#plotDat_Outside_degronvsWithin_degronSGS <- vst(dds)[sigGenes_Outside_degronvsWithin_degronSGS,] %>% 
#  assay()
plotDat_Outside_degronvsWithin_degronSGS <- vsd[sigGenes_Outside_degronvsWithin_degronSGS,] %>% 
  assay()
z.mat_Outside_degronvsWithin_degronSGS <- t(scale(t(plotDat_Outside_degronvsWithin_degronSGS), center=TRUE, scale=TRUE))

pdf(file = paste0(deseq_outDir, "/heatmaps/heatmap_top300DEgenes_Outside_degronvsWithin_degronSGS_batchEffRemoved.pdf"),
    width = 12, height = 8)
Heatmap(z.mat_Outside_degronvsWithin_degronSGS, name = "z-score",
        col = myRamp,            
        show_row_name = FALSE,
        #split=3,
        rect_gp = gpar(col = "lightgrey", lwd=0.3),
        top_annotation = ha)
dev.off()

# Get top 300 differentially expressed genes between Within_degron vs. truncating

res6$geneIDs = rownames(res6)

sigGenes_within_degronSGSvstruncating = as.data.frame(res6) %>% 
                                   top_n(300, wt=-padj) %>% 
                                   pull("geneIDs")

#plotDat_Outside_degronvsWithin_degronSGS <- vst(dds)[sigGenes_within_degronSGSvstruncating,] %>% 
#  assay()
plotDat_within_degronSGSvstruncating <- vsd[sigGenes_within_degronSGSvstruncating,] %>% 
  assay()
z.mat_within_degronSGSvstruncating <- t(scale(t(plotDat_within_degronSGSvstruncating), center=TRUE, scale=TRUE))

pdf(file = paste0(deseq_outDir, "/heatmaps/heatmap_top300DEgenes_within_degronSGSvstruncating_batchEffRemoved.pdf"),
    width = 12, height = 8)
Heatmap(z.mat_within_degronSGSvstruncating, name = "z-score",
        col = myRamp,            
        show_row_name = FALSE,
        #split=3,
        rect_gp = gpar(col = "lightgrey", lwd=0.3),
        top_annotation = ha)
dev.off()

# Get top 300 differentially expressed genes between Outside_degron vs. truncating

res7$geneIDs = rownames(res7)

sigGenes_Outside_degronvstruncating = as.data.frame(res7) %>% 
                                   top_n(300, wt=-padj) %>% 
                                   pull("geneIDs")

#plotDat_Outside_degronvsWithin_degronSGS <- vst(dds)[sigGenes_Outside_degronvsWithin_degronSGS,] %>% 
#  assay()
plotDat_Outside_degronvstruncating <- vsd[sigGenes_Outside_degronvstruncating,] %>% 
  assay()
z.mat_Outside_degronvstruncating <- t(scale(t(plotDat_Outside_degronvstruncating), center=TRUE, scale=TRUE))

pdf(file = paste0(deseq_outDir, "/heatmaps/heatmap_top300DEgenes_Outside_degronvstruncating_batchEffRemoved.pdf"),
    width = 12, height = 8)
Heatmap(z.mat_Outside_degronvstruncating, name = "z-score",
        col = myRamp,            
        show_row_name = FALSE,
        #split=3,
        rect_gp = gpar(col = "lightgrey", lwd=0.3),
        top_annotation = ha)
dev.off()

```

```{r goanalysis}

# tmp_df = data.frame(rn = res1$symbol, res1) %>%
#        na.omit() %>%
#        mutate(up = ifelse(log2FoldChange>0, padj, 1))

# only for upregulated genes
for (i in 1:7) {
      
  if (i==4) next
  # tmp_df = data.frame(rn = rownames(get(paste0("res",i))), get(paste0("res",i))) %>%
  #      na.omit() %>%
  #      mutate(up = ifelse(log2FoldChange>0, padj, 1))
  
  tmp_df = data.frame(rn = get(paste0("res",i))$symbol, get(paste0("res",i))) %>%
       na.omit() %>%
       mutate(up = ifelse(log2FoldChange>0, padj, 1))

# set the p-value for down regulations to 1
# make a named vector
tmp_condition_up = tmp_df$up
names(tmp_condition_up) <- tmp_df$rn

topDiffGenes<-function(x){ return(x<=0.1) }

# 1) BP

tg.BP = new("topGOdata",#just a name
            description="tmp_condition_up",
            ontology="BP",
            allGenes=tmp_condition_up, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.BP, paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_BP_upRegulated_res", i,".rds"))

numGenes(tg.BP)

numSigGenes(tg.BP)

fisher.res = runTest(tg.BP, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_BP = GenTable(tg.BP, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# 2) CC

tg.CC = new("topGOdata",#just a name
            description="tmp_condition_up",
            ontology="CC",
            allGenes=tmp_condition_up, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.CC, paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_CC_upRegulated_res", i,".rds"))

numGenes(tg.CC)

numSigGenes(tg.CC)

fisher.res = runTest(tg.CC, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_CC = GenTable(tg.CC, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# 3) MF

tg.MF = new("topGOdata",#just a name
            description="tmp_condition_up",
            ontology="MF",
            allGenes=tmp_condition_up, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.MF, paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_MF_upRegulated_res", i,".rds"))

numGenes(tg.MF)

numSigGenes(tg.MF)

fisher.res = runTest(tg.MF, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_MF = GenTable(tg.MF, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Plot

fishtab_BP2 = fishtab_BP[1:10, c(2,4)]
fishtab_BP2$ontology = "BP"
fishtab_CC2 = fishtab_CC[1:10, c(2,4)]
fishtab_CC2$ontology = "CC"
fishtab_MF2 = fishtab_MF[1:10, c(2,4)]
fishtab_MF2$ontology = "MF"

p1 = ggplot(fishtab_BP2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "green") + ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())
p2 = ggplot(fishtab_CC2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "red") + labs(x = "GO Terms") +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() +
  theme(axis.title.x=element_blank())
p3 = ggplot(fishtab_MF2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "blue") +  labs(y = "Number of genes") +  ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.title.y=element_blank())

plot_grid(p1, p2, p3, nrow = 3, ncol = 1)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO/GOtermsUp_res", i,".pdf"), width = 8, height = 15)

go_termsList_BP = fishtab_BP[1:10,] %>% arrange(desc(Significant))
write.table(go_termsList_BP, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_termList_UpBP_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_CC = fishtab_CC[1:10, ]%>% arrange(desc(Significant))
write.table(go_termsList_CC, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_termList_UpCC_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_MF = fishtab_MF[1:10, ]%>% arrange(desc(Significant))
write.table(go_termsList_MF, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_termList_UpMF_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")
}


# for downregulated genes

for (i in 1:7) {
      
  if (i==4) next
  # tmp_df = data.frame(rn = rownames(get(paste0("res",i))), get(paste0("res",i))) %>%
  #      na.omit() %>%
  #      mutate(down = ifelse(log2FoldChange<0, padj, 1))
  tmp_df = data.frame(rn = get(paste0("res",i))$symbol, get(paste0("res",i))) %>%
       na.omit() %>%
       mutate(down = ifelse(log2FoldChange<0, padj, 1))
# set the p-value for down regulations to 1
# make a named vector
tmp_condition_down = tmp_df$down
names(tmp_condition_down) <- tmp_df$rn

topDiffGenes<-function(x){ return(x<=0.1) }

# 1) BP

tg.BP = new("topGOdata",#just a name
            description="tmp_condition_down",
            ontology="BP",
            allGenes=tmp_condition_down, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.BP, paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_BP_downRegulated_res", i,".rds"))

numGenes(tg.BP)

numSigGenes(tg.BP)

fisher.res = runTest(tg.BP, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_BP = GenTable(tg.BP, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# retrieve geens2GO list from the annoation in GO data
  allGO_BP = genesInTerm(tg.BP)
  
# 2) CC

tg.CC = new("topGOdata",#just a name
            description="tmp_condition_down",
            ontology="CC",
            allGenes=tmp_condition_down, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.CC, paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_CC_downRegulated_res", i,".rds"))

numGenes(tg.CC)

numSigGenes(tg.CC)

fisher.res = runTest(tg.CC, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_CC = GenTable(tg.CC, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])
# retrieve geens2GO list from the annoation in GO data
  allGO_CC = genesInTerm(tg.CC)

# 3) MF

tg.MF = new("topGOdata",#just a name
            description="tmp_condition_down",
            ontology="MF",
            allGenes=tmp_condition_down, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.MF, paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_MF_downRegulated_res", i,".rds"))

numGenes(tg.MF)

numSigGenes(tg.MF)

fisher.res = runTest(tg.MF, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_MF = GenTable(tg.MF, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# retrieve geens2GO list from the annoation in GO data
  allGO_MF = genesInTerm(tg.MF)

  # Plot

fishtab_BP2 = fishtab_BP[1:10, c(2,4)]
fishtab_BP2$ontology = "BP"
fishtab_CC2 = fishtab_CC[1:10, c(2,4)]
fishtab_CC2$ontology = "CC"
fishtab_MF2 = fishtab_MF[1:10, c(2,4)]
fishtab_MF2$ontology = "MF"

p1 = ggplot(fishtab_BP2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "green") + ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())
p2 = ggplot(fishtab_CC2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "red") + labs(x = "GO Terms") +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() +
  theme(axis.title.x=element_blank())
p3 = ggplot(fishtab_MF2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "blue") +  labs(y = "Number of genes") +  ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.title.y=element_blank())

plot_grid(p1, p2, p3, nrow = 3, ncol = 1)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO/GOtermsDown_res", i,".pdf"), width = 8, height = 15)

go_termsList_BP = fishtab_BP[1:10,] %>% arrange(desc(Significant))
write.table(go_termsList_BP, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_termList_downBP_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_CC = fishtab_CC[1:10, ]%>% arrange(desc(Significant))
write.table(go_termsList_CC, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_termList_downCC_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_MF = fishtab_MF[1:10, ]%>% arrange(desc(Significant))
write.table(go_termsList_MF, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO/GO_termList_downMF_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")
}

#####this part doesn't work anymore so ignored, need to check orginal script#####
  # Make gene lists for each GO term in BP, CC and MF, for every condition pair
  
  go_terms_BP = fishtab_BP$GO.ID[1:10]
  go_terms_CC = fishtab_CC$GO.ID[1:10]
  go_terms_MF = fishtab_MF$GO.ID[1:10]
  
  for (j in c("BP", "CC", "MF")) {
    
    tmp_annotation = lapply(get(paste0("allGO_", j)), function(x) x[x %in% names(tmp_condition_up)])
    
      for (l in 1:10) {
        
        tmp_ensembl_gene_list = tmp_annotation[[get(paste0("go_terms_", j))[l]]]
        
        geneIDs = ensembldb::select(EnsDb.Hsapiens.v79, 
                                    keys = tmp_ensembl_gene_list, 
                                    keytype = "GENEID", 
                                    columns = c("SYMBOL","GENEID"))
        write.table(geneIDs, 
                    paste0(deseq_outDir, "GOtermAnalysis/topGO/gene_lists/GO_geneList_", j, "_",
                           noquote(get(paste0("go_terms_", j))[l]), 
                           "_res_", i, ".txt"), quote = F, row.names = F)
        
      }
  }
#}


```
# GOterm analysis
## topGO

```{r goanalysis: 1 UP: topGo ok}
# first follow what Gokberk did using topGO
# only for upregulated genes


# Function to retrieve full GO term names
get_full_GO_terms <- function(go_ids) {
 sapply(go_ids, function(go_id) {
   tryCatch({
     Term(GOTERM[[go_id]])
   }, error = function(e) {
     return(NA)
   })
 })
}

# Perform GO term enrichment and plotting
for (i in 1:7) {
      
  if (i==4) next
  # tmp_df = data.frame(rn = rownames(get(paste0("res",i))), get(paste0("res",i))) %>%
  #      na.omit() %>%
  #      mutate(up = ifelse(log2FoldChange>0, padj, 1))
  tmp_df = data.frame(rn = get(paste0("res",i))$symbol, get(paste0("res",i))) %>%
       na.omit() %>%
       mutate(up = ifelse(log2FoldChange>0, padj, 1))
library
# set the p-value for up regulations to 1
# make a named vector
tmp_condition_up = tmp_df$up
names(tmp_condition_up) <- tmp_df$rn

topDiffGenes<-function(x){ return(x<=0.1) }

# 1) BP

tg.BP = new("topGOdata",#just a name
            description="tmp_condition_up",
            ontology="BP",
            allGenes=tmp_condition_up, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL", #need to choose correct gene annotation for org.Hs.eg.db
            mapping="org.Hs.eg.db")

saveRDS(tg.BP, paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_BP_upRegulated_res", i,".rds"))

numGenes(tg.BP)

numSigGenes(tg.BP)

fisher.res = runTest(tg.BP, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_BP = GenTable(tg.BP, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Retrieve full GO terms for BP
fishtab_BP$Term = get_full_GO_terms(fishtab_BP$GO.ID)

# 2) CC

tg.CC = new("topGOdata",#just a name
            description="tmp_condition_up",
            ontology="CC",
            allGenes=tmp_condition_up, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.CC, paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_CC_upRegulated_res", i,".rds"))

numGenes(tg.CC)

numSigGenes(tg.CC)

fisher.res = runTest(tg.CC, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_CC = GenTable(tg.CC, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Retrieve full GO terms for CC
fishtab_CC$Term = get_full_GO_terms(fishtab_CC$GO.ID)


# 3) MF

tg.MF = new("topGOdata",#just a name
            description="tmp_condition_up",
            ontology="MF",
            allGenes=tmp_condition_up, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.MF, paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_MF_upRegulated_res", i,".rds"))

numGenes(tg.MF)

numSigGenes(tg.MF)

fisher.res = runTest(tg.MF, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_MF = GenTable(tg.MF, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Retrieve full GO terms for MF
fishtab_MF$Term = get_full_GO_terms(fishtab_MF$GO.ID)

# save top 30 GO term list
# The "significant" column indicates the number of genes or proteins significantly associated with each GO term. It provides a direct measure of the strength of association between the GO term and the gene/protein set, accounting for multiple testing if appropriate.
go_termsList_BP = fishtab_BP[1:30,] %>% arrange(desc(Significant))
write.table(go_termsList_BP, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_termList_top30upBP_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_CC = fishtab_CC[1:30, ]%>% arrange(desc(Significant))
write.table(go_termsList_CC, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_termList_top30upCC_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_MF = fishtab_MF[1:30, ]%>% arrange(desc(Significant))
write.table(go_termsList_MF, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_termList_top30upMF_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# Extract and save genes in each GO term for BP, CC, and MF
  genes_BP <- genesInTerm(tg.BP, fishtab_BP$GO.ID[1:20])
  genes_CC <- genesInTerm(tg.CC, fishtab_CC$GO.ID[1:20])
  genes_MF <- genesInTerm(tg.MF, fishtab_MF$GO.ID[1:20])
  
# Save to text file
  sink(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_terms_genes_up_res", i, ".txt"))
  cat("BP Genes:\n")
  for (go_id in names(genes_BP)) {
    cat(paste("GO Term:", go_id, "\n"))
    cat(paste(genes_BP[[go_id]], collapse = "\n"), "\n\n")
  }
  cat("CC Genes:\n")
  for (go_id in names(genes_CC)) {
    cat(paste("GO Term:", go_id, "\n"))
    cat(paste(genes_CC[[go_id]], collapse = "\n"), "\n\n")
  }
  cat("MF Genes:\n")
  for (go_id in names(genes_MF)) {
    cat(paste("GO Term:", go_id, "\n"))
    cat(paste(genes_MF[[go_id]], collapse = "\n"), "\n\n")
  }
  sink()

# Plot without REVIGO

fishtab_BP2 = fishtab_BP[1:10, c(2,4)]
fishtab_BP2$ontology = "BP"
fishtab_CC2 = fishtab_CC[1:10, c(2,4)]
fishtab_CC2$ontology = "CC"
fishtab_MF2 = fishtab_MF[1:10, c(2,4)]
fishtab_MF2$ontology = "MF"

# Wrap text for y-axis labels
  fishtab_BP2$Term <- str_wrap(fishtab_BP2$Term, width = 40)
  fishtab_CC2$Term <- str_wrap(fishtab_CC2$Term, width = 40)
  fishtab_MF2$Term <- str_wrap(fishtab_MF2$Term, width = 40)

p1 = ggplot(fishtab_BP2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "darkgreen") + ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.text.y = element_text(size = 8),
    axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 4), "cm"))
p2 = ggplot(fishtab_CC2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "purple") + labs(x = "GO Terms") +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() +
  theme(axis.text.y = element_text(size = 8),
    axis.title.x=element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 4), "cm"))
p3 = ggplot(fishtab_MF2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "orange") +  labs(y = "Number of genes") +  ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.text.y = element_text(size = 8),
    axis.title.y=element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 4), "cm"))

# # dont know why not saved probably when ru in the for loop? but ok separately
# #pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GOtermsup_res", i,".pdf"), width = 8, height = 15)
plot_grid(p1, p2, p3, nrow = 3, ncol = 1)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GOtermsup_res", i,".pdf"), width = 8, height = 15)
# #dev.off()

# use rrvgo to reduce redundant GO terms https://ssayols.github.io/rrvgo/articles/rrvgo.html
# Get reduced redundant GO terms for all GO terms sorted by Fisher pvalue

# BP
go_termsList_BP = fishtab_BP[1:50, ] #%>% arrange(desc(Significant))
#go_termsList_BP = fishtab_BP
simMatrix_BP <- calculateSimMatrix(go_termsList_BP$GO.ID,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")
# Convert the "Fisher" column to numeric
go_termsList_BP$Fisher <- as.numeric(go_termsList_BP$Fisher)
scores_BP <- setNames(-log10(go_termsList_BP$Fisher), go_termsList_BP$GO.ID)
reducedTerms_BP <- reduceSimMatrix(simMatrix_BP,
                                scores_BP,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
#save top50 reduced GO terms
write.table(reducedTerms_BP,
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGO_top50termList_upBP_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# plot relationship of GO terms usinf scatter plots
#pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50upBP_scatter_res", i,".pdf"), width = 8, height = 8)
scatterPlot(simMatrix_BP, reducedTerms_BP)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50upBP_scatter_res", i,".pdf"), width = 8, height = 8)
#dev.off()

# CC
go_termsList_CC = fishtab_CC[1:50, ] #%>% arrange(desc(Significant))
#go_termsList_CC = fishtab_CC
simMatrix_CC <- calculateSimMatrix(go_termsList_CC$GO.ID,
                                 orgdb="org.Hs.eg.db",
                                 ont="CC",
                                 method="Rel")
# Convert the "Fisher" column to numeric
go_termsList_CC$Fisher <- as.numeric(go_termsList_CC$Fisher)
scores_CC <- setNames(-log10(go_termsList_CC$Fisher), go_termsList_CC$GO.ID)
reducedTerms_CC <- reduceSimMatrix(simMatrix_CC,
                                 scores_CC,
                                 threshold=0.7,
                                 orgdb="org.Hs.eg.db")
#save top50 reduced GO terms
write.table(reducedTerms_CC,
               paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGO_termList_top50upCC_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# plot relationship of GO terms usinf scatter plots
# pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50upCC_scatter_res", i,".pdf"), width = 8, height = 8)
scatterPlot(simMatrix_CC, reducedTerms_CC)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50upCC_scatter_res", i,".pdf"), width = 8, height = 8)
# dev.off()

# MF
go_termsList_MF = fishtab_MF[1:50, ] #%>% arrange(desc(Significant))
#go_termsList_MF = fishtab_MF
simMatrix_MF <- calculateSimMatrix(go_termsList_MF$GO.ID,
                                 orgdb="org.Hs.eg.db",
                                 ont="MF",
                                 method="Rel")
# Convert the "Fisher" column to numeric
go_termsList_MF$Fisher <- as.numeric(go_termsList_MF$Fisher)
scores_MF <- setNames(-log10(go_termsList_MF$Fisher), go_termsList_MF$GO.ID)
reducedTerms_MF <- reduceSimMatrix(simMatrix_MF,
                                 scores_MF,
                                 threshold=0.7,
                                 orgdb="org.Hs.eg.db")
#save top50 reduced GO terms
write.table(reducedTerms_MF,
               paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGO_termList_top50upMF_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# plot relationship of GO terms usinf scatter plots
# pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50upMF_scatter_res", i,".pdf"), width = 8, height = 8)
scatterPlot(simMatrix_MF, reducedTerms_MF)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50upMF_scatter_res", i,".pdf"), width = 8, height = 8)

# dev.off()
}


```

DID'NT RUN FOR RES5-7

```{r goanalysis: 1 DOWN: topGo ok}
# first follow what Gokberk did using topGO
# only for downregulated genes


# Function to retrieve full GO term names
get_full_GO_terms <- function(go_ids) {
 sapply(go_ids, function(go_id) {
   tryCatch({
     Term(GOTERM[[go_id]])
   }, error = function(e) {
     return(NA)
   })
 })
}

# Perform GO term enrichment and plotting
for (i in 1:7) {
      
  if (i==4) next
  # tmp_df = data.frame(rn = rownames(get(paste0("res",i))), get(paste0("res",i))) %>%
  #      na.omit() %>%
  #      mutate(down = ifelse(log2FoldChange<0, padj, 1))
  tmp_df = data.frame(rn = get(paste0("res",i))$symbol, get(paste0("res",i))) %>%
       na.omit() %>%
       mutate(down = ifelse(log2FoldChange<0, padj, 1))
library
# set the p-value for down regulations to 1
# make a named vector
tmp_condition_down = tmp_df$down
names(tmp_condition_down) <- tmp_df$rn

topDiffGenes<-function(x){ return(x<=0.1) }

# 1) BP

tg.BP = new("topGOdata",#just a name
            description="tmp_condition_down",
            ontology="BP",
            allGenes=tmp_condition_down, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL", #need to choose correct gene annotation for org.Hs.eg.db
            mapping="org.Hs.eg.db")

saveRDS(tg.BP, paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_BP_downRegulated_res", i,".rds"))

numGenes(tg.BP)

numSigGenes(tg.BP)

fisher.res = runTest(tg.BP, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_BP = GenTable(tg.BP, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Retrieve full GO terms for BP
fishtab_BP$Term = get_full_GO_terms(fishtab_BP$GO.ID)

# 2) CC

tg.CC = new("topGOdata",#just a name
            description="tmp_condition_down",
            ontology="CC",
            allGenes=tmp_condition_down, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.CC, paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_CC_downRegulated_res", i,".rds"))

numGenes(tg.CC)

numSigGenes(tg.CC)

fisher.res = runTest(tg.CC, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_CC = GenTable(tg.CC, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Retrieve full GO terms for CC
fishtab_CC$Term = get_full_GO_terms(fishtab_CC$GO.ID)


# 3) MF

tg.MF = new("topGOdata",#just a name
            description="tmp_condition_down",
            ontology="MF",
            allGenes=tmp_condition_down, #named vector with adj. p-values
            geneSel=topDiffGenes,
            annot=annFUN.org,
            nodeSize = 10, #minimum number of genes in a GO categorie
            ID="SYMBOL",
            mapping="org.Hs.eg.db")

saveRDS(tg.MF, paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_MF_downRegulated_res", i,".rds"))

numGenes(tg.MF)

numSigGenes(tg.MF)

fisher.res = runTest(tg.MF, algorithm="elim", statistic="fisher")
fisher.res

#p-value adjustment
qval = p.adjust(score(fisher.res), method = "BH")
summary(qval<0.1)

fishtab_MF = GenTable(tg.MF, Fisher=fisher.res, orderBy="Fisher", topNodes = geneData(fisher.res)[4])

# Retrieve full GO terms for MF
fishtab_MF$Term = get_full_GO_terms(fishtab_MF$GO.ID)

# save top 30 GO term list
# The "significant" column indicates the number of genes or proteins significantly associated with each GO term. It provides a direct measure of the strength of association between the GO term and the gene/protein set, accounting for multiple testing if appropriate.
go_termsList_BP = fishtab_BP[1:30,] %>% arrange(desc(Significant))
write.table(go_termsList_BP, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_termList_top30downBP_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_CC = fishtab_CC[1:30, ]%>% arrange(desc(Significant))
write.table(go_termsList_CC, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_termList_top30downCC_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")


go_termsList_MF = fishtab_MF[1:30, ]%>% arrange(desc(Significant))
write.table(go_termsList_MF, 
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_termList_top30downMF_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# Extract and save genes in each GO term for BP, CC, and MF
  genes_BP <- genesInTerm(tg.BP, fishtab_BP$GO.ID[1:20])
  genes_CC <- genesInTerm(tg.CC, fishtab_CC$GO.ID[1:20])
  genes_MF <- genesInTerm(tg.MF, fishtab_MF$GO.ID[1:20])
  
# Save to text file
  sink(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GO_terms_genes_down_res", i, ".txt"))
  cat("BP Genes:\n")
  for (go_id in names(genes_BP)) {
    cat(paste("GO Term:", go_id, "\n"))
    cat(paste(genes_BP[[go_id]], collapse = "\n"), "\n\n")
  }
  cat("CC Genes:\n")
  for (go_id in names(genes_CC)) {
    cat(paste("GO Term:", go_id, "\n"))
    cat(paste(genes_CC[[go_id]], collapse = "\n"), "\n\n")
  }
  cat("MF Genes:\n")
  for (go_id in names(genes_MF)) {
    cat(paste("GO Term:", go_id, "\n"))
    cat(paste(genes_MF[[go_id]], collapse = "\n"), "\n\n")
  }
  sink()

# Plot without REVIGO

fishtab_BP2 = fishtab_BP[1:10, c(2,4)]
fishtab_BP2$ontology = "BP"
fishtab_CC2 = fishtab_CC[1:10, c(2,4)]
fishtab_CC2$ontology = "CC"
fishtab_MF2 = fishtab_MF[1:10, c(2,4)]
fishtab_MF2$ontology = "MF"

# Wrap text for y-axis labels
  fishtab_BP2$Term <- str_wrap(fishtab_BP2$Term, width = 40)
  fishtab_CC2$Term <- str_wrap(fishtab_CC2$Term, width = 40)
  fishtab_MF2$Term <- str_wrap(fishtab_MF2$Term, width = 40)

p1 = ggplot(fishtab_BP2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "darkgreen") + ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.text.y = element_text(size = 8),
    axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 4), "cm"))
p2 = ggplot(fishtab_CC2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "purple") + labs(x = "GO Terms") +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() +
  theme(axis.text.y = element_text(size = 8),
    axis.title.x=element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 4), "cm"))
p3 = ggplot(fishtab_MF2, aes(x = reorder(Term, Significant), y = Significant)) +
  geom_bar(stat = "identity", fill = "orange") +  labs(y = "Number of genes") +  ylim(0, max(fishtab_CC2$Significant)) +
  coord_flip() + facet_wrap(~ontology) + theme_minimal() + 
  theme(axis.text.y = element_text(size = 8),
    axis.title.y=element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 4), "cm"))

# # dont know why not saved probably when ru in the for loop? but ok separately
# #pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GOtermsDown_res", i,".pdf"), width = 8, height = 15)
plot_grid(p1, p2, p3, nrow = 3, ncol = 1)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/GOtermsDown_res", i,".pdf"), width = 8, height = 15)
# #dev.off()

# use rrvgo to reduce redundant GO terms https://ssayols.github.io/rrvgo/articles/rrvgo.html
# Get reduced redundant GO terms for all GO terms sorted by Fisher pvalue

# BP
go_termsList_BP = fishtab_BP[1:50, ] #%>% arrange(desc(Significant))
#go_termsList_BP = fishtab_BP
simMatrix_BP <- calculateSimMatrix(go_termsList_BP$GO.ID,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")
# Convert the "Fisher" column to numeric
go_termsList_BP$Fisher <- as.numeric(go_termsList_BP$Fisher)
scores_BP <- setNames(-log10(go_termsList_BP$Fisher), go_termsList_BP$GO.ID)
reducedTerms_BP <- reduceSimMatrix(simMatrix_BP,
                                scores_BP,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
#save top50 reduced GO terms
write.table(reducedTerms_BP,
              paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGO_top50termList_downBP_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# plot relationship of GO terms usinf scatter plots
#pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50DownBP_scatter_res", i,".pdf"), width = 8, height = 8)
scatterPlot(simMatrix_BP, reducedTerms_BP)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50DownBP_scatter_res", i,".pdf"), width = 8, height = 8)
#dev.off()

# CC
go_termsList_CC = fishtab_CC[1:50, ] #%>% arrange(desc(Significant))
#go_termsList_CC = fishtab_CC
simMatrix_CC <- calculateSimMatrix(go_termsList_CC$GO.ID,
                                 orgdb="org.Hs.eg.db",
                                 ont="CC",
                                 method="Rel")
# Convert the "Fisher" column to numeric
go_termsList_CC$Fisher <- as.numeric(go_termsList_CC$Fisher)
scores_CC <- setNames(-log10(go_termsList_CC$Fisher), go_termsList_CC$GO.ID)
reducedTerms_CC <- reduceSimMatrix(simMatrix_CC,
                                 scores_CC,
                                 threshold=0.7,
                                 orgdb="org.Hs.eg.db")
#save top50 reduced GO terms
write.table(reducedTerms_CC,
               paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGO_termList_top50downCC_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# plot relationship of GO terms usinf scatter plots
# pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50DownCC_scatter_res", i,".pdf"), width = 8, height = 8)
scatterPlot(simMatrix_CC, reducedTerms_CC)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50DownCC_scatter_res", i,".pdf"), width = 8, height = 8)
# dev.off()

# MF
go_termsList_MF = fishtab_MF[1:50, ] #%>% arrange(desc(Significant))
#go_termsList_MF = fishtab_MF
simMatrix_MF <- calculateSimMatrix(go_termsList_MF$GO.ID,
                                 orgdb="org.Hs.eg.db",
                                 ont="MF",
                                 method="Rel")
# Convert the "Fisher" column to numeric
go_termsList_MF$Fisher <- as.numeric(go_termsList_MF$Fisher)
scores_MF <- setNames(-log10(go_termsList_MF$Fisher), go_termsList_MF$GO.ID)
reducedTerms_MF <- reduceSimMatrix(simMatrix_MF,
                                 scores_MF,
                                 threshold=0.7,
                                 orgdb="org.Hs.eg.db")
#save top50 reduced GO terms
write.table(reducedTerms_MF,
               paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGO_termList_top50downMF_res_", i, ".txt"), quote = F, row.names = F, sep = "\t")

# plot relationship of GO terms usinf scatter plots
# pdf(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50DownMF_scatter_res", i,".pdf"), width = 8, height = 8)
scatterPlot(simMatrix_MF, reducedTerms_MF)
ggsave(paste0(deseq_outDir, "GOtermAnalysis/topGO_MW/reducedGOterms_top50DownMF_scatter_res", i,".pdf"), width = 8, height = 8)

# dev.off()
}


```



```{r vennDiagram}
#deseq_outDir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_HDF/diffexpression/MW_20221018/results/DESeq/",sep="")

# Outside degron vs within degron vs truncating

# Significantly UPregulated genes in Control vs 3 conditions
# outside_up = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigUpRegulatedGenes_res1.tab"), header = T)
# inside_up = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigUpRegulatedGenes_res2.tab"), header = T)
# truncating_up = read.table(paste0(deseq_outDir, "sigDEgeneLists/sigUpRegulatedGenes_res3.tab"), header = T)

outside_up = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res1_sigUpRegulatedGenes.tab"), header = T)
outside_up = na.omit(outside_up)
nrow(outside_up)
write.csv(outside_up, file =paste0(deseq_outDir, "sigDEgeneListsMW/res1_sigUpRegulatedGenes_noNA.csv"))

inside_up = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res2_sigUpRegulatedGenes.tab"), header = T)
inside_up = na.omit(inside_up)
nrow(inside_up)
write.csv(inside_up, file =paste0(deseq_outDir, "sigDEgeneListsMW/res2_sigUpRegulatedGenes_noNA.csv"))

truncating_up = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res3_sigUpRegulatedGenes.tab"), header = T)
truncating_up = na.omit(truncating_up)
nrow(truncating_up)
write.csv(truncating_up, file =paste0(deseq_outDir, "sigDEgeneListsMW/res3_sigUpRegulatedGenes_noNA.csv"))

n12_up = length(intersect(rownames(outside_up),rownames(inside_up)))
write.table(outside_up[intersect(rownames(outside_up),rownames(inside_up)),"symbol"],
            paste0(deseq_outDir, "/sigDEgeneListsMW/sigUpRegulatedGenes_outsidevsInsideOverlap_geneList.tab"),
            quote = F, row.names = F, col.names = F)
n23_up = length(intersect(rownames(inside_up),rownames(truncating_up)))
write.table(outside_up[intersect(rownames(inside_up),rownames(truncating_up)),"symbol"],
            paste0(deseq_outDir, "/sigDEgeneListsMW/sigUpRegulatedGenes_insidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)
n13_up = length(intersect(rownames(outside_up),rownames(truncating_up)))
write.table(outside_up[intersect(rownames(outside_up),rownames(truncating_up)),"symbol"],
            paste0(deseq_outDir, "/sigDEgeneListsMW/sigUpRegulatedGenes_outsidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)
n123_up = length(intersect(rownames(truncating_up),intersect(rownames(outside_up),rownames(inside_up))))
write.table(outside_up[intersect(rownames(truncating_up),intersect(rownames(outside_up),rownames(inside_up))),"symbol"],
            paste0(deseq_outDir, "/sigDEgeneListsMW/sigUpRegulatedGenes_truncatingvsOutsidevsInside_geneList.tab"),
            quote = F, row.names = F, col.names = F)

venn_up = draw.triple.venn(nrow(outside_up),nrow(inside_up),nrow(truncating_up),
                 n12_up, n23_up, n13_up, n123_up,
                 c("Outside_degron", "Within_degron", "Truncating"),
                 fill = c("deepskyblue", "orange", "darkorchid1"),
                 lwd = 0,  # Removes the black border
                  cat.cex = 1.5,  # Adjusts category label size
                  cex = 1.2,  # Adjusts the number size
                  fontface = "plain",  # Ensures a consistent font style
                  cat.fontface = "plain",
                  cat.fontfamily = "sans",  # Change category labels to Arial
                  fontfamily = "sans"  # Change numbers to Arial
)

pdf(paste0(deseq_outDir, "vennDiagrams/GA/venn_sigDEgenesUp_outsideVSinsideVStruncating.pdf"),
    width = 8, height = 8)
grid.draw(venn_up)
dev.off()

# Significantly DOWNregulated genes in Control vs 3 conditions
outside_down = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res1_sigDownRegulatedGenes.tab"), header = T)
outside_down = na.omit(outside_down)
nrow(outside_down)
write.csv(outside_down, file =paste0(deseq_outDir, "sigDEgeneListsMW/res1_sigDownRegulatedGenes_noNA.csv"))

inside_down = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res2_sigDownRegulatedGenes.tab"), header = T)
inside_down = na.omit(inside_down)
nrow(inside_down)
write.csv(inside_down, file =paste0(deseq_outDir, "sigDEgeneListsMW/res2_sigDownRegulatedGenes_noNA.csv"))

truncating_down = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res3_sigDownRegulatedGenes.tab"), header = T)
truncating_down = na.omit(truncating_down)
nrow(truncating_down)
write.csv(truncating_down, file =paste0(deseq_outDir, "sigDEgeneListsMW/res3_sigDownRegulatedGenes_noNA.csv"))

n12_down = length(intersect(rownames(outside_down),rownames(inside_down)))
write.table(outside_down[intersect(rownames(outside_down),rownames(inside_down)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDownRegulatedGenes_outsidevsInsideOverlap_geneList.tab"),
            quote = F, row.names = F, col.names = F)
n23_down = length(intersect(rownames(inside_down),rownames(truncating_down)))
write.table(outside_down[intersect(rownames(inside_down),rownames(truncating_down)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDownRegulatedGenes_insidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)
n13_down = length(intersect(rownames(outside_down),rownames(truncating_down)))
write.table(outside_down[intersect(rownames(outside_down),rownames(truncating_down)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDownRegulatedGenes_outsidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)
n123_down = length(intersect(rownames(truncating_down),intersect(rownames(outside_down),rownames(inside_down))))
write.table(outside_down[intersect(rownames(truncating_down),intersect(rownames(outside_down),rownames(inside_down))),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDownRegulatedGenes_truncatingvsOutsidevsInside_geneList.tab"),
            quote = F, row.names = F, col.names = F)

venn_down = draw.triple.venn(nrow(outside_down),nrow(inside_down),nrow(truncating_down),
                 n12_down, n23_down, n13_down, n123_down,
                 c("Outside_degron", "Within_degron", "Truncating"),
                 fill = c("deepskyblue", "orange", "darkorchid1"),
                 lwd = 0,  # Removes the black border
                  cat.cex = 1.5,  # Adjusts category label size
                  cex = 1.2,  # Adjusts the number size
                  fontface = "plain",  # Ensures a consistent font style
                  cat.fontface = "plain",
                  cat.fontfamily = "sans",  # Change category labels to Arial
                  fontfamily = "sans"  # Change numbers to Arial
)
pdf(paste0(deseq_outDir, "vennDiagrams/GA/venn_sigDEgenesDown_outsideVSinsideVStruncating.pdf"),
    width = 8, height = 8)
grid.draw(venn_down)
dev.off()

# Make a Venn Diagram for all sig DE genes
# and also a print out list of them
# done

outside_DEG = rbind(outside_up, outside_down)
nrow(outside_DEG)
write.csv(outside_DEG, file =paste0(deseq_outDir, "sigDEgeneLists/res1_sigAllDEGenes_noNA.csv"))

inside_DEG = rbind(inside_up, inside_down)
nrow(inside_DEG)
write.csv(inside_DEG, file =paste0(deseq_outDir, "sigDEgeneLists/res2_sigAllDEGenes_noNA.csv"))

truncating_DEG = rbind(truncating_up, truncating_down)
nrow(truncating_DEG)
write.csv(truncating_DEG, file =paste0(deseq_outDir, "sigDEgeneLists/res3_sigAllDEGenes_noNA.csv"))


n12_DEG = length(intersect(rownames(outside_DEG),rownames(inside_DEG)))
write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(inside_DEG)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneLists/sigDEG_outsidevsInsideOverlap_geneList.tab"),
            quote = F, row.names = F, col.names = F)

n23_DEG = length(intersect(rownames(inside_DEG),rownames(truncating_DEG)))
write.table(outside_DEG[intersect(rownames(inside_DEG),rownames(truncating_DEG)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneLists/sigDEG_insidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)

n13_DEG = length(intersect(rownames(outside_DEG),rownames(truncating_DEG)))
write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(truncating_DEG)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneLists/sigDEG_outsidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)

n123_DEG = length(intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))))
write.table(outside_DEG[intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))),"symbol"],
            paste0(deseq_outDir, "sigDEgeneLists/sigDEG_truncatingvsOutsidevsInside_geneList.tab"),
            quote = F, row.names = F, col.names = F)

venn_DEG = draw.triple.venn(nrow(outside_DEG),nrow(inside_DEG),nrow(truncating_DEG),
                 n12_DEG, n23_DEG, n13_DEG, n123_DEG,
                 c("Outside_degron", "Within_degron", "Truncating"),
                 fill = c("deepskyblue", "orange", "darkorchid1"),
                 lwd = 0,  # Removes the black border
                  cat.cex = 1.5,  # Adjusts category label size
                  cex = 1.2,  # Adjusts the number size
                  fontface = "plain",  # Ensures a consistent font style
                  cat.fontface = "plain",
                  cat.fontfamily = "sans",  # Change category labels to Arial
                  fontfamily = "sans"  # Change numbers to Arial
)
pdf(paste0(deseq_outDir, "vennDiagrams/GA/venn_sigDEG_outsideVSinsideVStruncating_v1.pdf"),
    width = 8, height = 8)
grid.draw(venn_DEG)
dev.off()

# Make a Venn Diagram for all sig DE genes (from mine)
outside_DEG = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res1_sigAllDEGenes.tab"), header = T)
outside_DEG = na.omit(outside_DEG)
nrow(outside_DEG)
write.csv(outside_DEG, file =paste0(deseq_outDir, "sigDEgeneListsMW/res1_sigAllDEGenes_noNA.csv"))

inside_DEG = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res2_sigAllDEGenes.tab"), header = T)
inside_DEG = na.omit(inside_DEG)
nrow(inside_DEG)
write.csv(inside_DEG, file =paste0(deseq_outDir, "sigDEgeneListsMW/res2_sigAllDEGenes_noNA.csv"))

truncating_DEG = read.table(paste0(deseq_outDir, "sigDEgeneListsMW/res3_sigAllDEGenes.tab"), header = T)
truncating_DEG = na.omit(truncating_DEG)
nrow(truncating_DEG)
write.csv(truncating_DEG, file =paste0(deseq_outDir, "sigDEgeneListsMW/res3_sigAllDEGenes_noNA.csv"))


n12_DEG = length(intersect(rownames(outside_DEG),rownames(inside_DEG)))
write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(inside_DEG)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDEG_outsidevsInsideOverlap_geneList.tab"),
            quote = F, row.names = F, col.names = F)

n23_DEG = length(intersect(rownames(inside_DEG),rownames(truncating_DEG)))
write.table(outside_DEG[intersect(rownames(inside_DEG),rownames(truncating_DEG)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDEG_insidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)

n13_DEG = length(intersect(rownames(outside_DEG),rownames(truncating_DEG)))
write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(truncating_DEG)),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDEG_outsidevsTruncating_geneList.tab"),
            quote = F, row.names = F, col.names = F)

n123_DEG = length(intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))))
write.table(outside_DEG[intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))),"symbol"],
            paste0(deseq_outDir, "sigDEgeneListsMW/sigDEG_truncatingvsOutsidevsInside_geneList.tab"),
            quote = F, row.names = F, col.names = F)

venn_DEG = draw.triple.venn(nrow(outside_DEG),nrow(inside_DEG),nrow(truncating_DEG),
                 n12_DEG, n23_DEG, n13_DEG, n123_DEG,
                 c("Outside_degron", "Within_degron", "Truncating"),
                 fill = c("deepskyblue", "orange", "darkorchid1"),
                 lwd = 0,  # Removes the black border
                  cat.cex = 1.5,  # Adjusts category label size
                  cex = 1.2,  # Adjusts the number size
                  fontface = "plain",  # Ensures a consistent font style
                  cat.fontface = "plain",
                  cat.fontfamily = "sans",  # Change category labels to Arial
                  fontfamily = "sans"  # Change numbers to Arial
)
pdf(paste0(deseq_outDir, "vennDiagrams/GA/venn_sigDEG_outsideVSinsideVStruncating_v2.pdf"),
    width = 8, height = 8)
grid.draw(venn_DEG)
dev.off()

# # Significantly DEGs in Control vs 3 conditions
# outside_DEG = read.csv(paste0(deseq_outDir, "deseq_results/res_ControlvsOutsideDegron.csv"), header = T)
# outside_DEG = subset(outside_DEG, outside_DEG$padj<0.1)
# outside_DEG = subset(outside_DEG, outside_DEG$log2FoldChange >= 1 | outside_DEG$log2FoldChange <= -1)
# outside_DEG = data.frame(outside_DEG)
# rownames(outside_DEG)=outside_DEG$X
# outside_DEG = outside_DEG[,-1]
# nrow(outside_DEG)
# #[1] 208
# write.table(outside_DEG, paste0(deseq_outDir,"/sigDEgeneLists/filteredDEG_ControlvsOutside.tab"),
#               quote = F, sep = "\t")
# #removeNA
# outside_DEG = read.table(paste0(deseq_outDir, "/sigDEgeneLists/filteredDEG_ControlvsOutside.tab"), header = T)
# outside_DEG_noNA = na.omit(outside_DEG)
# nrow(outside_DEG_noNA)
# #[1] 197
# write.table(outside_DEG_noNA, paste0(deseq_outDir,"/sigDEgeneLists/filteredDEG_ControlvsOutside_noNA.tab"),
#               quote = F, sep = "\t")
# 
# 
# 
# inside_DEG = read.csv(paste0(deseq_outDir, "deseq_results/res_ControlvsWithin_degronSGS.csv"), header = T)
# inside_DEG = subset(inside_DEG, inside_DEG$padj<0.1)
# inside_DEG = subset(inside_DEG, inside_DEG$log2FoldChange >= 1 | inside_DEG$log2FoldChange <= -1)
# inside_DEG = data.frame(inside_DEG)
# rownames(inside_DEG)=inside_DEG$X
# inside_DEG = inside_DEG[,-1]
# nrow(inside_DEG)
# #[1] 389
# write.table(inside_DEG, paste0(deseq_outDir,"/sigDEgeneLists/filteredDEG_ControlvsWithin_degronSGS.tab"),
#               quote = F, sep = "\t")
# #removeNA
# inside_DEG = read.table(paste0(deseq_outDir, "/sigDEgeneLists/filteredDEG_ControlvsWithin_degronSGS.tab"), header = T)
# inside_DEG_noNA = na.omit(inside_DEG)
# nrow(inside_DEG_noNA)
# #[1] 366
# write.table(inside_DEG_noNA, paste0(deseq_outDir,"/sigDEgeneLists/filteredDEG_ControlvsWithin_degronSGS_noNA.tab"),
#               quote = F, sep = "\t")
# 
# 
# truncating_DEG = read.csv(paste0(deseq_outDir, "deseq_results/res_ControlvsTruncating.csv"), header = T)
# truncating_DEG = subset(truncating_DEG, truncating_DEG$padj<0.1)
# truncating_DEG = subset(truncating_DEG, truncating_DEG$log2FoldChange >= 1 | truncating_DEG$log2FoldChange <= -1)
# truncating_DEG = data.frame(truncating_DEG)
# rownames(truncating_DEG)=truncating_DEG$X
# truncating_DEG = truncating_DEG[,-1]
# nrow(truncating_DEG)
# #[1] 367
# write.table(truncating_DEG, paste0(deseq_outDir,"/sigDEgeneLists/filteredDEG_ControlvsTruncating.tab"),
#               quote = F, sep = "\t")
# 
# #removeNA
# truncating_DEG = read.table(paste0(deseq_outDir, "/sigDEgeneLists/filteredDEG_ControlvsTruncating.tab"), header = T)
# truncating_DEG_noNA = na.omit(truncating_DEG)
# nrow(truncating_DEG_noNA)
# #[1] 345
# write.table(truncating_DEG_noNA, paste0(deseq_outDir,"/sigDEgeneLists/filteredDEG_ControlvsTruncating_noNA.tab"),
#               quote = F, sep = "\t")
# 
# 
# n12_DEG = length(intersect(rownames(outside_DEG),rownames(inside_DEG)))
# write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(inside_DEG)),"symbol"],
#             paste0(deseq_outDir, "/sigDEgeneListsMW/DEG_outsidevsInsideOverlap_geneList.tab"),
#             quote = F, row.names = F, col.names = F)
# n23_DEG = length(intersect(rownames(inside_DEG),rownames(truncating_DEG)))
# write.table(outside_DEG[intersect(rownames(inside_DEG),rownames(truncating_DEG)),"symbol"],
#             paste0(deseq_outDir, "/sigDEgeneListsMW/DEG_insidevsTruncating_geneList.tab"),
#             quote = F, row.names = F, col.names = F)
# n13_DEG = length(intersect(rownames(outside_DEG),rownames(truncating_DEG)))
# write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(truncating_DEG)),"symbol"],
#             paste0(deseq_outDir, "/sigDEgeneListsMW/DEG_outsidevsTruncating_geneList.tab"),
#             quote = F, row.names = F, col.names = F)
# n123_DEG = length(intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))))
# write.table(outside_DEG[intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))),"symbol"],
#             paste0(deseq_outDir, "/sigDEgeneListsMW/DEG_truncatingvsOutsidevsInside_geneList.tab"),
#             quote = F, row.names = F, col.names = F)
# 
# venn_DEG = draw.triple.venn(nrow(outside_DEG),nrow(inside_DEG),nrow(truncating_DEG),
#                  n12_DEG, n23_DEG, n13_DEG, n123_DEG,
#                  c("Outside_degron", "Inside_degron", "Truncating"),
#                  fill = c("blue", "red", "green"))
# pdf(paste0(deseq_outDir, "vennDiagrams/GA/venn_DEG_outsideVSinsideVStruncating.pdf"),
#     width = 12, height = 8)
# grid.draw(venn_DEG)
# dev.off()

# #for overlapping of DEGs, autism and ID genes-----------------here--> this is odd, do i need this?start here
# #outside
# outsideDEG_ASD = length(intersect(rownames(outside_DEG),rownames(autism_genes)))
# outsideDEG_ID = length(intersect(rownames(outside_DEG),rownames(ID_genes)))
# 
# truncatingDEG_ASD = length(intersect(rownames(autism_genes),rownames(truncating_DEG)))
# truncatingDEG_ID = length(intersect(rownames(ID_genes),rownames(truncating_DEG)))
# # 0 overlap for ASD or ID genes with truncating only
# insideDEG_ASD = length(intersect(rownames(inside_DEG),rownames(autism_genes)))
# insideDEG_ID = length(intersect(rownames(inside_DEG),rownames(ID_genes)))
# # 0 overlap for ASD or ID genes with inside only
# 
# 115
# write.table(outside_DEG[intersect(rownames(outside_DEG),rownames(truncating_DEG)),"symbol"],
#             paste0(deseq_outDir, "/sigDEgeneListsMW/DEG_outsidevsTruncating_geneList.tab"),
#             quote = F, row.names = F, col.names = F)
# n123_DEG = length(intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))))
# write.table(outside_DEG[intersect(rownames(truncating_DEG),intersect(rownames(outside_DEG),rownames(inside_DEG))),"symbol"],
#             paste0(deseq_outDir, "/sigDEgeneListsMW/DEG_truncatingvsOutsidevsInside_geneList.tab"),
#             quote = F, row.names = F, col.names = F)
# 
# venn_DEG = draw.triple.venn(nrow(outside_DEG),nrow(inside_DEG),nrow(truncating_DEG),
#                  n12_DEG, n23_DEG, n13_DEG, n123_DEG,
#                  c("Outside_degron", "Inside_degron", "Truncating"),
#                  fill = c("blue", "red", "green"))
# pdf(paste0(deseq_outDir, "vennDiagrams/GA/venn_DEG_outsideVSinsideVStruncating.pdf"),
#     width = 12, height = 8)
# grid.draw(venn_DEG)
# dev.off()



```


# RRHO analysis to check global transcriptomics similarity between conditions

```{r RRHO-2}
#https://systems.crump.ucla.edu/rankrank/index.php
# do for all genes
# make lists for RRHO and save them as csv files, this works
for (i in 1:7) {
  if (i==4) next
  
  tmp_dds = get(paste0("res", i))
  
  # remove genes which could not be evaluated
  # tmp_subset = data.frame(tmp_dds,
  #                         stringsAsFactors = F %>%
  #                         )
  
#  tmp_dds = tmp_dds[tmp_dds$padj<=0.05,] # here we get only sig DEs 
  tmp_list = data.frame(element = tmp_dds$symbol,
                        value = -log10(tmp_dds$pvalue)*sign(tmp_dds$log2FoldChange))
  
  write.csv(tmp_list, 
            file = paste0(deseq_outDir, "RRHO_allGenes/list_", i, ".csv"),
            quote = F, row.names = F)
  
}

# run RRHO for each list pair, only do for res 1, 2, 3: Res1: control vs outside, Res2: control vs within SGS, Res3: control vs truncating LoF
# now it also works
  for (j in  1:3) {
    if (j==4) next
    
    tmp_list1 = read.csv(paste0(deseq_outDir, "RRHO_allGenes/list_", j, ".csv"))
    tmp_list1 = na.omit(tmp_list1) #remove the rows without a gene symbol
    tmp_list1 = tmp_list1[order(tmp_list1$element, -abs(tmp_list1$value)),] #remove the duplicate genes and keep the one with a larger DE
    tmp_list1 = tmp_list1[!duplicated(tmp_list1$element),]
    
    for (k in  1:3) {
      if (k==4|k==j) next
      
      tmp_list2 = read.csv(paste0(deseq_outDir, "RRHO_allGenes/list_", k, ".csv"))
      tmp_list2 = na.omit(tmp_list2)
      tmp_list2 = tmp_list2[order(tmp_list2$element, -abs(tmp_list2$value)),]
      tmp_list2 = tmp_list2[!duplicated(tmp_list2$element),]
      
  RRHO=    RRHO(tmp_list1, tmp_list2, 
           stepsize = 100, 
           labels = (c(as.character(j),as.character(k))), 
           alternative = "enrichment", #100 us recommended as the step size
           plots = T, outputdir = paste0(deseq_outDir, "RRHO_allGenes/"), 
           BY = T,
           log10.ind = T) 
#same gene list length so no need to correct for that      

              }
  }



```

## plot normalized data for specific genes of interests

```{r boxplot2grid}
# Load necessary libraries
# library(DESeq2)
# library(dplyr)
# library(ggplot2)
# library(gridExtra)

# dds4 here is only for plotting boxplots since i want symbols but not ensembl ids

# dds4 = dds3
# convert_to_symbol <- function(dds4) {
#   symbols <- mapIds(org.Hs.eg.db,
#                     keys = rownames(dds4),
#                     column = "SYMBOL",
#                     keytype = "ENSEMBL",
#                     multiVals = "first")
#   rownames(dds4) <- symbols
#   return(dds4)
# }
# 
# dds4 <- convert_to_symbol(dds4)

# dds4 <- dds4[!is.na(rownames(dds4)), ]
# nrow(dds4)
# #[1] 14682

# Read genes of interest
goi = read.csv(paste0(dir, "lg-scrna/working_data/Maggie/working/setbp1_LoForg_bulkrna_b1/GOI_ok.csv"), header = TRUE)
colnames(goi) = c("gene", "remarks")
goi = goi$gene

for (i in 1:3) {
  # Load DESeq2 result object
  tmp_dds = get(paste0("res", i))

  # Remove genes that could not be evaluated
  tmp_subset = data.frame(
    #gene = rownames(tmp_dds),
    tmp_dds,
    stringsAsFactors = FALSE
  ) %>%
    na.omit() %>%
    dplyr::filter(padj < 0.1) %>%
    arrange(desc(abs(log2FoldChange))) #%>%
    #rownames(tmp_subset) = tmp_subset$symbol
  
  # Get top 10 DEGs
  top_10_DEGs = list(head(tmp_subset$symbol, 20))[[1]]
  goi2 = c(top_10_DEGs, goi)
  
  total_genes = length(goi2)
  processed_genes = 0
  plots <- list()
  
  # Determine global y-axis limits
  all_counts <- c()
  
  for (j in goi2) {
    if (is.na(j)) {
      next
    }
    #if (!any(tmp_subset$gene == j) && !any(rownames(tmp_dds) == j)) {
    if (!any(tmp_subset$symbol == j, na.rm = TRUE) && !any(tmp_dds$symbol == j, na.rm = TRUE)) {
      next
    }
    
    tryCatch({
      counts_data <- plotCounts(dds4, gene = j, intgroup = "condition", returnData = TRUE)
      all_counts <- c(all_counts, counts_data$count)
    }, error = function(e) {
      cat("Error in extracting counts for gene", j, ":", e$message, "\n")
    })
  }
  
  y_min <- min(all_counts, na.rm = TRUE)
  y_max <- max(all_counts, na.rm = TRUE)
  
  for (j in goi2) {
    if (is.na(j)) {
      cat("Skipping NA gene\n")
      next
    }
    
    #if (!any(tmp_subset$gene == j) && !any(rownames(tmp_dds) == j)) {
    if (!any(tmp_subset$symbol == j, na.rm = TRUE) && !any(tmp_dds$symbol == j, na.rm = TRUE)) {

      cat("Gene", j, "not in filtered tmp_subset or tmp_dds. Attempting to plot from original dataset.\n")
      
      # Attempt to plot from original dataset if data is available
      if (!any(rownames(dds4) == j)) {
        cat("Gene", j, "not found in original dataset, skipping\n")
        next
      }
    }
    
    cat("Processing gene", j, "\n")
    processed_genes <- processed_genes + 1
    
    tryCatch({
      # Extract counts for the gene
      counts_data <- plotCounts(dds4, gene = j, intgroup = "condition", returnData = TRUE)
      
      # Set the order of condition
      counts_data$condition <- factor(counts_data$condition, levels = c("Control", "Within_degron", "Outside_degron", "Truncating"))
      
      # set colour
      condition_colors <- c("Control" = "azure4",
                            "Within_degron" = "orange",
                      "Outside_degron" = "deepskyblue",
                      "Truncating" = "darkorchid1")

  
      # Create a box plot
      p <- ggplot(counts_data, aes(x = condition, y = count, fill = condition)) +
        geom_boxplot(alpha = 0.5) +
        geom_jitter(position = position_jitter(width = 0.1), color = "black", alpha = 0.5) +  # Add jittered points
        labs(title = j, x = "Condition", y = "Log10(Normalized Count)") +
        theme_classic() +
        scale_y_log10(limits = c(y_min, y_max)) +
        scale_fill_manual(values = condition_colors) + # Apply custom colors
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
      
      # Add the plot to the list of plots
      plots[[length(plots) + 1]] <- p
    }, error = function(e) {
      cat("Error in plotting gene", j, ":", e$message, "\n")
    })
  }
  
  # Arrange plots in a grid and save to a single PDF
  if (length(plots) > 0) {
    nrow_plots <- ceiling(length(plots) / 4)
    pdf(file = paste0(deseq_outDir, "gene_boxplots/all_goi/res", i, "_combined.pdf"), width = 16, height = 4 * nrow_plots)
    grid.arrange(grobs = plots, ncol = 4)
    dev.off()
  }
  
  cat("Total genes in goi2:", total_genes, "\n")
  cat("Total processed genes:", processed_genes, "\n")
}

```

### this find overlaps between two gene lists, do fisher exact test & plot venn diagram

```{r overlaps 3}
# Load necessary libraries
# library(dplyr)
# library(VennDiagram)
# library(gridExtra)
# library(stats)  # for Fisher's exact test

# Define a function to plot Venn diagram and save as PDF
plot_venn_pdf <- function(list_of_genes, labels, output_file) {
  pdf(file = output_file, width = 10, height = 10)
  venn.plot <- venn.diagram(
    x = list_of_genes,
    category.names = labels,
    filename = NULL,  # Do not save as a file/.png
    output = TRUE,
    col = "black",
    fill = c("cornflowerblue", "green"),  # Adjusted to 2 colors for 2 sets
    alpha = 0.50,
    cex = 2,
    fontface = "bold",
    fontfamily = "sans",
    cat.cex = 2,
    cat.fontface = "bold",
    cat.default.pos = "outer",
    cat.pos = c(-27, 27),  # Adjusted for 2 sets
    cat.dist = c(0.055, 0.055),  # Adjusted for 2 sets
    cat.fontfamily = "sans",
    rotation.degree = 1  # Changed from rotation to rotation.degree
  )
  grid.draw(venn.plot)
  dev.off()
}

# Define a function to perform Fisher's exact test
perform_fishers_test <- function(set1, set2, universe_size) {
  overlap <- length(intersect(set1, set2))
  only_set1 <- length(set1) - overlap
  only_set2 <- length(set2) - overlap
  neither <- universe_size - length(union(set1, set2))
  
  contingency_table <- matrix(c(overlap, only_set1, only_set2, neither), 
                              nrow = 2, 
                              dimnames = list(c("In Set2", "Not in Set2"),
                                              c("In Set1", "Not in Set1")))
  
  fisher_test_result <- fisher.test(contingency_table)
  return(fisher_test_result$p.value)
}

# Read genes of interest
overlapList_Dir = paste(dir, "lg-scrna/working_data/Maggie/working/bulk_rna_SETBP1_iNeuron/",sep="")

setbp1_targets <- read.csv(paste0(overlapList_Dir, "Top70_putative_SETBP1_bindingsites.csv"), header = F)$V1
setbp1_targets[1] = "EP300"
ID_genes <- read.csv(paste0(overlapList_Dir, "ID_whole_panelapp.csv"), header = F)$V1
autism_genes <- read.csv2(paste0(overlapList_Dir, "autism_whole_panelapp.csv"), header = F)$V1
ectopicG870S <- read.csv2(paste0(overlapList_Dir, "ectopicG870S.csv"), header = F)$V1 #this one somehow has a lot of tabs in front of the gene name

DEG_LoF_HDF <- read.csv(paste0(overlapList_Dir, "res_ControlvsTruncating_noNA.csv"), header = T)  %>%
    dplyr::filter(padj < 0.1)
DEG_LoF_HDF <- DEG_LoF_HDF$symbol

DEG_SGS_HDF <- read.csv(paste0(overlapList_Dir, "res_ControlvsWithin_degronSGS_noNA.csv"), header = T)  %>%
    dplyr::filter(padj < 0.1)
DEG_SGS_HDF <- DEG_SGS_HDF$symbol

DEG_outside_HDF <- read.csv(paste0(overlapList_Dir, "res_ControlvsOutsideDegron_noNA.csv"), header = T)  %>%
    dplyr::filter(padj < 0.1)
DEG_outside_HDF <- DEG_outside_HDF$symbol

# Get DESeq files
deseq_files <- list.files(paste0(deseq_outDir, "sigDEgeneListsMW"), pattern = "_sigAllDEGenes.csv", full.names = T)

# Function to clean gene names
clean_gene_names <- function(gene_vector) {
  return(trimws(toupper(gene_vector)))
}

# Clean gene names in input lists
setbp1_targets <- clean_gene_names(setbp1_targets)
ID_genes <- clean_gene_names(ID_genes)
autism_genes <- clean_gene_names(autism_genes)
ectopicG870S <- clean_gene_names(ectopicG870S) #now there are no more tabs in front of gene names

# Define the universe size (this should be the total number of genes considered in your DESeq analysis)
universe_size <- 14481  # Replace with your actual universe size, got this from nrow(res1_deseq_results_noNA)

# Check overlap for ID & autism genes
overlap_ID_autism <- intersect(ID_genes, autism_genes)
if (length(overlap_ID_autism) > 0) {
  write.csv(overlap_ID_autism, 
            file = paste0(deseq_outDir, "overlaps/ID_autism_overlap.csv"),
            quote = F, row.names = F)
  
  plot_venn_pdf(
    list(ID_genes, autism_genes),
    c("ID Genes", "Autism Genes"),
    paste0(deseq_outDir, "overlaps/ID_autism_venn.pdf")
  )
  
  p_value <- perform_fishers_test(ID_genes, autism_genes, universe_size)
  write.csv(data.frame(p_value = p_value), 
            file = paste0(deseq_outDir, "overlaps/ID_autism_fishers_pvalue.csv"),
            quote = F, row.names = F)
} else {
  print(paste("No overlap for ID & autism genes"))
}

# Process DESeq files
for (i in deseq_files) {
  
  tmp_f <- read.csv(i, header = TRUE)
  tmp_f_name <- tail(strsplit(i, "/")[[1]], 1) # always fetch the last part of the split path, which is the file name
  
  # Clean gene names in DESeq result
  tmp_f$X <- clean_gene_names(tmp_f$symbol)
  
  # Check overlap for setbp1 targets
  overlap_setbp1 <- intersect(setbp1_targets, tmp_f$X)
  if (length(overlap_setbp1) > 0) {
    write.csv(overlap_setbp1, 
              file = paste0(deseq_outDir, "overlaps/setbp1Targets_overlap_", tmp_f_name),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(setbp1_targets, tmp_f$X),
      c("SETBP1 Targets", tmp_f_name),
      paste0(deseq_outDir, "overlaps/setbp1Targets_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(setbp1_targets, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/setbp1Targets_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for setbp1 targets in file", tmp_f_name))
  }
  
  # Check overlap for ID genes
  overlap_ID_genes <- intersect(ID_genes, tmp_f$X)
  if (length(overlap_ID_genes) > 0) {
    write.csv(overlap_ID_genes, 
              file = paste0(deseq_outDir, "overlaps/panelapp_IDgenes_overlap_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(ID_genes, tmp_f$X),
      c("ID Genes", tmp_f_name),
      paste0(deseq_outDir, "overlaps/ID_genes_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(ID_genes, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/ID_genes_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for ID genes in file", tmp_f_name))
  }
  
  # Check overlap for autism genes
  overlap_autism_genes <- intersect(autism_genes, tmp_f$X)
  if (length(overlap_autism_genes) > 0) {
    write.csv(overlap_autism_genes, 
              file = paste0(deseq_outDir, "overlaps/panelapp_autism_genes_overlap_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(autism_genes, tmp_f$X),
      c("Autism Genes", tmp_f_name),
      paste0(deseq_outDir, "overlaps/autism_genes_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(autism_genes, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/autism_genes_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for autism genes in file", tmp_f_name))
  }
  
  # Check overlap for ectopicG870S
  overlap_ectopicG870S <- intersect(ectopicG870S, tmp_f$X)
  if (length(overlap_ectopicG870S) > 0) {
    write.csv(overlap_ectopicG870S, 
              file = paste0(deseq_outDir, "overlaps/ectopicG870S_DEG_overlap_", tmp_f_name),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(ectopicG870S, tmp_f$X),
      c("Ectopic G870S", tmp_f_name),
      paste0(deseq_outDir, "overlaps/ectopicG870S_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(ectopicG870S, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/ectopicG870S_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for ectopicG870S in file", tmp_f_name))
  }
  
  # Check overlap for DEG of SETBP1 LoF HDF
  overlap_DEG_LoF_HDF <- intersect(DEG_LoF_HDF, tmp_f$X)
  if (length(overlap_DEG_LoF_HDF) > 0) {
    write.csv(overlap_DEG_LoF_HDF, 
              file = paste0(deseq_outDir, "overlaps/DEG_LoF_HDF_overlap_", tmp_f_name),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(DEG_LoF_HDF, tmp_f$X),
      c("DEG LoF HDF", tmp_f_name),
      paste0(deseq_outDir, "overlaps/DEG_LoF_HDF_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(DEG_LoF_HDF, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/DEG_LoF_HDF_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for DEG_LoF_HDF in file", tmp_f_name))
  }
  
  # Check overlap for DEG of SETBP1 SGS HDF
  overlap_DEG_SGS_HDF <- intersect(DEG_SGS_HDF, tmp_f$X)
  if (length(overlap_DEG_SGS_HDF) > 0) {
    write.csv(overlap_DEG_SGS_HDF, 
              file = paste0(deseq_outDir, "overlaps/DEG_SGS_HDF_overlap_", tmp_f_name),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(DEG_SGS_HDF, tmp_f$X),
      c("DEG SGS HDF", tmp_f_name),
      paste0(deseq_outDir, "overlaps/DEG_SGS_HDF_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(DEG_SGS_HDF, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/DEG_SGS_HDF_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for DEG_SGS_HDF in file", tmp_f_name))
  }
  
  
  # Check overlap for DEG of SETBP1 SGS HDF
  overlap_DEG_outside_HDF <- intersect(DEG_outside_HDF, tmp_f$X)
  if (length(overlap_DEG_outside_HDF) > 0) {
    write.csv(overlap_DEG_outside_HDF, 
              file = paste0(deseq_outDir, "overlaps/DEG_outside_HDF_overlap_", tmp_f_name),
              quote = F, row.names = F)
    
    plot_venn_pdf(
      list(DEG_outside_HDF, tmp_f$X),
      c("DEG outside HDF", tmp_f_name),
      paste0(deseq_outDir, "overlaps/DEG_outside_HDF_venn_", tmp_f_name, ".pdf")
    )
    
    p_value <- perform_fishers_test(DEG_outside_HDF, tmp_f$X, universe_size)
    write.csv(data.frame(p_value = p_value), 
              file = paste0(deseq_outDir, "overlaps/DEG_outside_HDF_fishers_pvalue_", tmp_f_name, ".csv"),
              quote = F, row.names = F)
  } else {
    print(paste("No overlap for DEG_outside_HDF in file", tmp_f_name))
  }
  
  summary(tmp_f$X)
}

```


```{r session info}
sessionInfo()

```

